---
title: "Hotdog Module Answer Key"
format: html
---

Data Summary:

1.  Why might it be difficult to compare earlier winners to most recent winners?

    The total times of the competitions vary between years so we can not compare the total hotdogs eaten of different years.

2.  Create a standard variable to compare winners across time by dividing the hotdog variable by the time variable to get "Hotdogs_per_minute".

    ```{r}
    library(readr)
    library(tidyverse)
    hotdog_clean <- read_csv("hotdogs/hotdog_clean.csv")
    hotdog_men <- hotdog_clean |> filter(Sex == 1)
    hotdog_women <- hotdog_clean |> filter(Sex == 0)

    hotdog_men_min <- hotdog_men |> mutate("Hotdogs_per_minute" = Dogs/Time)

    hotdog_women_min <-hotdog_women |> mutate("Hotdogs_per_minute" = Dogs/Time)
    ```

3.  Create a scatterplot of the Amount of Hotdogs consumed per minute by year.

    ```{r}
    ggplot(data = hotdog_men_min, aes(x = Year, y = Hotdogs_per_minute)) +
      geom_point() +
      theme_minimal()

    ggplot(data = hotdog_women_min, aes(x = Year, y = Hotdogs_per_minute)) +
      geom_point() +
      theme_minimal()
    ```

4.  What do you notice about the amount of hotdogs consumed per minute in 2001?

    There is a big jump in number of hotdogs consumed per minute between 2000 and 2001.

5.  Looking at the original "Dogs" variable, what is a possible explanation for the trend shown on the plot?

    In earlier years the total time for the competition varied between 2-6 minutes so competitors did not eat as many hotdogs but ate a greater amount proportionally per minute.

6.  Who was the winner in 2001?

    Takeru Kobyashi
