---
title: "Studenet copy"
format: html
---

Module: Tour de France — Data, Distance, and Dynamics
Course: Introduction to Data Science
Format: Follow-Along / Homework
Author: [Your Name]

🎯 Introduction & Learning Objectives
The Tour de France is a grueling, multi-day, multi-stage bike race around France and surrounding countries. Athletes are pushed to their absolute physical and psychological limits for ego, earnings, teammates, and national pride. (Fun fact: The famous White Jersey is awarded to the best rider under 26!).

In this module, you will learn how to:

Import data directly from the web into R.

Wrangle and clean data using dplyr (filtering, mutating).

Create and manipulate different types of graphs using ggplot2.

Build and interpret an introductory linear regression model.

🛠️ Part 1: Getting the Data
First, we need to load our libraries and import our dataset. Instead of downloading files to our computers, we can pull data directly from the internet using the TidyTuesday project, which hosts fantastic datasets for learning.

Code snippet
# Load our core packages
```{r}
library(tidyverse)

# Fetch the data directly from the TidyTuesday GitHub repository
tdf_winners <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-04-07/tdf_winners.csv')
```
```{r}
# Let's take a quick look at what we are working with
head(tdf_winners)
```
🚲 Part 2: Visualizing the Race Over Time
Let's start by looking at how the total distance of the Tour de France has changed across all its editions (years).

1. Distance over Time
Code snippet
```{r}
ggplot(data = ___, aes(x = ___, y = ___)) +
  geom_point(color = "darkblue") +
  labs(title = "Tour de France Total Distance by Edition",
       x = "Edition (Year)",
       y = "Distance (km)") +
  theme_minimal()
```
Student Prompt 1: Look at the graph above. Is there a trend? What is it? Based on this graph, what might we expect to see if we look at the total time riders spend racing, or their average speeds?

2. Calculating and Plotting Speed
The dataset gives us distance and time_overall, but it doesn't give us the speed. We can calculate this ourselves using the mutate() function. Speed is simply Distance divided by Time.

Code snippet
# Clean the data and calculate speed
```{r}
tdf_speed <- tdf_winners |>
  group_by(edition) |>
  ___(edition, distance, time_overall) |>  # Keep only these three columns
  ___(!is.na(distance), !is.na(time_overall)) |> # Remove missing values
  ___(speed = distance / time_overall) # Create the new speed column

# Check your work:
head(tdf_speed)
```

Code snippet
```{r}
# YOUR CODE HERE: Plot time_overall vs edition, coloring the points by time_overall

```
# Plot the new speed variable
```{r}
# YOUR CODE HERE: Create a scatter plot of speed vs. edition

```
⏱️ Part 3: Identifying Historical "Eras"
If we plot the average spped to win the race, we can start to see distinct groupings or "eras" in the data.

Student Prompt 2: Why do these graphs look the way they do? What historical world events might have caused gaps or massive shifts in the data during the early-to-mid 1900s?

Let's explicitly parse the data into time periods to see if our theory holds up. We will use case_when() to create a new column called era.

Code snippet
```{r}
tdfed <- tdf_speed |> 
  mutate(era = case_when(
    edition <= 21 ~ "1903 - 1924",
    edition <= 36 ~ "1925 - 1939",
    edition <= 60 ~ "1947 - 1971 (Post WWII)",
    edition <= 80 ~ "1972 - 1992",
    edition <= 106 ~ "1993 - 2019",
    TRUE ~ "2020 - Present"
  ))
```

# Plotting with our new 'era' variable
```{r}
ggplot(data = tdfed, aes(x = edition, y = speed)) +
  geom_point(aes(colour = era), size = 2) +
  labs(title = "TDF Speeds Grouped by Era",
       x = "Edition",
       y = "Speed (km/h)") +
  theme_minimal()
```
Student Prompt 3: Looking at the color groupings, why are there distinct periods where the speed dramatically jumps or plateaus? Consider advancements in technology, training, or race rules.

📈 Part 4: Intro to Regression Modeling
Now we are going to plot Distance and Speed on the same graph using a dual-axis, and try to fit a linear regression model to predict speed based on distance.

The First Model
Code snippet
```{r}
ggplot(tdf_speed, aes(x = edition)) +
  geom_line(aes(y = speed, color = "Speed")) +
  geom_point(aes(y = speed, color = "Speed")) +
  
  # We divide distance by 100 to scale it to fit nicely on the same graph as speed
  geom_line(aes(y = distance/100, color = "Distance")) +
  geom_point(aes(y = distance/100, color = "Distance")) +
  
  scale_y_continuous(
    name = "Speed (km/h)",
    sec.axis = sec_axis(~.*100, name = "Distance (km)")
  ) +
  labs(x = "Edition", color = "Metric", title = "TDF: Speed vs. Distance") +
  theme_minimal()
```
# Build a linear model: Does distance predict speed?
```{r}
# Build the linear model
mod_tdf1 = lm(___ ~ ___, data = ___)

# View the results
summary(___)
```
Student Prompt 4: Look at the R-squared and the p-values in the summary above. What is wrong with this model? Think about the early years of the Tour vs. the modern years.

Fixing the Data (Model 2)
The early years of the Tour were incredibly long and heavily skew our data. Let's filter the data to only look at years where the distance was reasonably modernized (e.g., shorter than the massive early routes).

Code snippet
# Filter for editions where the distance was 3000km or more (ignoring extreme modern short routes or anomalies)
```{r}
# YOUR CODE HERE: Create 'tdfer' by filtering for distance >= 3000

# YOUR CODE HERE: Create 'mod_tdf2' using the tdfer dataset and summarize it
```
# Run the model again on the filtered data
```{r}
mod_tdf2 = lm(speed ~ distance, data = tdfer)
summary(mod_tdf2)
```
Student Prompt 5 (Homework): Compare the summary of mod_tdf2 to mod_tdf1. Did the model improve? Will this model always be correct for predicting the speed of future races based on recent years? Why or why not? (Hint: Can bikes get infinitely faster? Is there a human biological limit?)
