---
title: "How Much Does 0.1 Seconds Matter? Time and Ranking in NCAA Men's 50 Freestyle"
author:
  - name: Kevin Euyoque
    email: kevin.euyoque04@gmail.com
    affiliations:
      - ref: slu
  - name: Ivan Ramler
    email: iramler@stlawu.edu
    affiliations:
      - ref: slu
affiliations:
  - id: slu
    name: St. Lawrence University
date: "May 5, 2026"
description: "A data visualization tutorial exploring how small differences in race time relate to rankings in NCAA men's 50 freestyle swimming."
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
categories:
  - Sports Analytics
  - Data Visualization
  - NCAA Swimming
  - R
---

## How Much Does a Fraction of a Second Matter?
### Exploring Performance in NCAA Swimming

In competitive swimming, races are often decided by fractions of a second. At the NCAA level, the difference between first place and fifth place can be extremely small sometimes less than half a second. Despite these tiny margins, the outcomes can be dramatically different.

So how much does a small difference in time actually matter?

In this module, we explore real NCAA swimming data to understand how race time relates to ranking in one event: the men’s 50 freestyle.


## Learning Objectives

- Understand how performance is measured in competitive swimming  
- Explore how small differences in swim time impact rank  
- Analyze the relationship between swim time and rank for one NCAA swimming event  
- Interpret a linear regression model using a realistic 0.1-second change  


## Focus Event: Men’s 50 Freestyle

For this analysis, we focus on one event: the **men’s 50 freestyle**.

This event is one of the shortest races in competitive swimming, which makes it ideal for studying how small time differences affect outcomes.

The main variables we will use are:

- **Swim time** (seconds)  
- **Rank** (finishing position)

Lower times are better, and lower ranks are also better.  
So we expect that as swim time increases, rank will also increase.



## Load and Prepare the Data

```{r}
library(tidyverse)
library(readr)
```

```{r}
swim_df <- read_csv("kevineuyoque/usa_swimming/SwimData.csv")

swim_50_free_men <- swim_df |>
  filter(Event == "50 Freestyle SCY Male", Sex == "M") |>
  mutate(Swim_Time = as.numeric(Swim_Time)) |>
  select(Event, Swimmer, Sex, Swim_Time, Rank) |>
  drop_na(Swim_Time, Rank)

swim_50_free_men
```

## Visualizing the Relationship

We begin with a scatterplot to explore the relationship between swim time and rank.

```{r}
ggplot(swim_50_free_men, aes(x = Swim_Time, y = Rank)) +
  geom_point(alpha = 0.6) +
  scale_x_continuous(n.breaks = 8) +
  labs(title = "Men's 50 Freestyle: Swim Time vs. Rank",
    subtitle = "Slower times tend to correspond to worse rankings",
    x = "Swim Time (seconds)",
    y = "Rank") +
  theme_minimal()
```

### What This First Graph Shows

This scatterplot shows a clear pattern: as swim time increases, rank also tends to increase.

In swimming, this means that slower times are connected to worse rankings. Since **rank 1 is the best**, moving upward in rank number means finishing lower in the standings.

This is important because the x-axis only changes by small amounts. Even a few tenths of a second can separate many swimmers.


## Adding a Regression Line

The scatterplot shows the relationship visually, but we can also add a regression line to summarize the trend.

```{r}
ggplot(swim_50_free_men, aes(x = Swim_Time, y = Rank)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(n.breaks = 8) +
  labs(title = "Men's 50 Freestyle: Regression Line",
    subtitle = "The line summarizes how rank changes as swim time increases",
    x = "Swim Time (seconds)",
    y = "Rank") +
  theme_minimal()
```
The upward trend means that swimmers with slower times usually have worse rankings.

This fits what we would expect in a race. However, the important part is not just that slower swimmers rank lower. The important part is how quickly rank changes when time changes.

In a short race like the 50 freestyle, even a small time difference can matter.


## Building a Linear Model

Next, we create a linear regression model using swim time to predict rank

```{r}
rank_model <- lm(Rank ~ Swim_Time, data = swim_50_free_men)

summary(rank_model)
```
## Interpreting the Model Output

From the regression results, the most important value is the **slope for Swim_Time**, which is:

- **89.33**

This tells us how rank changes as swim time increases


### What Does the Slope Mean?

A standard interpretation would say:

- A **1-second increase** in swim time is associated with about a **89.33 increase in rank**

However, this is not very realistic in this context.

In a 50 freestyle race, a full second is an extremely large difference. NCAA swimmers are already competing at a very high level, so changes are much smaller


## A More Realistic Interpretation: 0.1 Seconds

Instead, we interpret a **0.1-second change**, which is much more realistic in swimming.

```{r}
0.1 * 89.326
```

Because lower rank is better, this means a swimmer who is **0.1 seconds slower** is predicted to finish about **9 places worse** in the rankings.

This is a much more meaningful interpretation than using a full second. In elite sprint swimming, one second is a huge difference, but one tenth of a second is a realistic margin.


## Model Strength

The model also has an R-squared value of about **0.82**.

This means that about **82% of the variation in rank** is explained by swim time.

That is a strong relationship, which makes sense because rankings in swimming are directly connected to finishing times.

## Why This Matters

The main takeaway from this analysis is that small differences in time can have a large impact on competitive results.

In the men’s 50 freestyle:

- a 1-second change is too large to be a useful interpretation
- a 0.1-second change is more realistic
- a 0.1-second slower time is associated with about 9 places worse in ranking

This shows why starts, turns, reaction time, and finishing technique matter so much in sprint events. Even a tiny improvement can shift a swimmer several places.


## Student Check-In

Answer the following questions based on the regression model:

1. Why is a 1-second increase not the best interpretation for this event?
2. What does a 0.1-second increase in swim time mean in terms of rank?
3. Since lower rank is better, what does a positive slope tell us?
4. Why might small time improvements matter more in the 50 freestyle than in longer events?


## Conclusion

In NCAA swimming, fractions of a second are not just small details. They can be the difference between finishing near the top or falling several places behind.

For the men’s 50 freestyle, our model suggests that even a **0.1-second difference** is associated with a meaningful change in rank.

This makes the 50 freestyle a strong example of how data can help explain competitive performance in sports.

