---
title: "Code to Wrangle Open Powerlifting Data"
format: html
editor: visual
---

## Wrangling Open Powerlifting Data

```{r}
library(readr)
library(curl)
library(dplyr)
library(tidyr)
library(tidyverse)
library(stringr)
library(ggplot2)
```

This chunk downloads the zip file containing the openpowerlifting data from <https://openpowerlifting.gitlab.io/opl-csv/bulk-csv.html>

```{r}
# Step 1: Set up the URL
zip_url <- "https://openpowerlifting.gitlab.io/opl-csv/files/openpowerlifting-latest.zip"

# Step 2: Use curl_download to save it safely
tmp_zip <- tempfile(fileext = ".zip")
curl_download(zip_url, tmp_zip)

# Step 3: List contents
zip_files <- unzip(tmp_zip, list = TRUE)
# print(zip_files$Name) # can double check names of files with this

# Step 4: Find the CSV file
csv_path <- zip_files$Name[grepl("\\.csv$", zip_files$Name)]

# Step 5: Read the CSV directly from the ZIP and clean up files
openpowerlifting <- read_csv(unz(tmp_zip, csv_path))
unlink(tmp_zip)

# Step 6: Check if the entry worked
# should be at least 3.6 million rows and 42 columns (as of June 2025)
dim(openpowerlifting)

```

This chunk cleans up the full dataset. Please note that this is a large dataset and the code may take a few minutes to run.

```{r}
openpl_tidy <-
  openpowerlifting |>
    group_by(Name) |> # these three lines extract the most recent
    arrange(desc(Date), # event for each person
            .by_group = TRUE) |>
    slice(1) |>
    filter( # these help standardize the lifters included in the dataset
      (Tested == "Yes")  # those that were drug tests
      & (Event == "SBD")# only the main event
      & Equipment %in% c("Single-ply","Raw") # standard equipment
      & !(Place %in% c("DQ", "NS", "G", "DD")) # only those that completed their lifts
      ) |>
    drop_na(Age, Best3SquatKg, 
            Best3BenchKg, Best3DeadliftKg) # ensuring the main variables are complete cases
    
```

We now split the data into the two most prominent sexes included in the dataset. (Note, Sex == 'Mx' are excluded do to the very small sample size.)

Additionally, we take a random sample of size 500 from each Age class to ensure that (i) all age classes are approximately equally represented and (ii) that the dataset is small enough to use in a classroom setting.

Note that for very young or old age classes, if there are fewer than 500 lifters, we do not take a random sample of them and instead include all lifters for that class.

```{r}
set.seed(501) 
# Fun fact: This seed is used as it is the 
# all-time world record deadlift of 501 kg (1,105 lb), 
# achieved by Iceland's Hafþór Júlíus Björnsson.
# (aka "The Mountain" from the TV Series, "Game of Thrones")

# Male Podium Lifters
openpl_male<-
  openpl_tidy |>
  filter(Sex == "M") |>
  select(-Sex,
         -Squat1Kg, -Squat2Kg, -Squat3Kg, -Squat4Kg,
         -Bench1Kg, -Bench2Kg, -Bench3Kg, -Bench4Kg,
         -Deadlift1Kg, -Deadlift2Kg, -Deadlift3Kg, -Deadlift4Kg,
         ) |>
  ungroup() |>
  mutate(id = row_number()) |>
  rename(Deadlift = Best3DeadliftKg,
         Bench = Best3BenchKg,
         Squat = Best3SquatKg
         ) |>
  relocate(id, .before = Name) |>
  # the following code takes the random sample
  group_by(AgeClass) %>%
  group_modify(~ if (nrow(.x) >= 20) sample_n(.x, 20) else .x) %>%
  ungroup() |>
  # keeping primary varaibles of interest
  select(Name,Age, AgeClass, BodyweightKg, 
         Squat, Bench, Deadlift, 
         TotalKg, Place, Dots, Wilks, Glossbrenner, Goodlift,
         Date)|>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClassInd = if_else(AgeClass == "80-84" |
                                 AgeClass == "85-89" |
                                 AgeClass == "90-999",
                               true = 1,
                               false = 0)) |>
  filter(AgeClassInd == 0) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12"))

```

```{r}
set.seed(290) 
# Fun fact: This seed is used as it is the 
# all-time female world record deadlift of 290 kg (639.4 lbs)
# achieved by America's Tamara Walcott

# Female Lifters
openpl_female<-
  openpl_tidy |>
  filter(Sex == "F") |>
  select(-Sex,
         -Squat1Kg, -Squat2Kg, -Squat3Kg, -Squat4Kg,
         -Bench1Kg, -Bench2Kg, -Bench3Kg, -Bench4Kg,
         -Deadlift1Kg, -Deadlift2Kg, -Deadlift3Kg, -Deadlift4Kg,
         ) |>
  ungroup() |>
  mutate(id = row_number()) |>
  rename(Deadlift = Best3DeadliftKg,
         Bench = Best3BenchKg,
         Squat = Best3SquatKg
         ) |>
  relocate(id, .before = Name) |>
  # the following code takes the random sample
  group_by(AgeClass) %>%
  group_modify(~ if (nrow(.x) >= 20) sample_n(.x, 20) else .x) %>%
  ungroup()  |>
  # keeping primary variables of interest
  select(Name, Age, AgeClass, BodyweightKg, 
         Squat, Bench, Deadlift, 
         TotalKg, Place, Dots, Wilks, Glossbrenner, Goodlift,
         Date) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClassInd = if_else(AgeClass == "80-84" |
                                 AgeClass == "85-89" |
                                 AgeClass == "90-999",
                               true = 1,
                               false = 0)) |>
  filter(AgeClassInd == 0) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12"))

```

```{r}
female_bench_ranked <- openpl_female |> arrange(desc(Bench)) |>
  mutate(rank = row_number()) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  select(Name, Age, AgeClass, Bench, rank)

female_bench_ranked |>
  group_by(AgeClass)  |>
  summarise(n())

ggplot(data = female_bench_ranked, aes(x = AgeClass, y = Bench)) +
  geom_boxplot() +
  theme_minimal()

female_bench_plot <- female_bench_ranked |>
  group_by(AgeClass)  |>
  summarise(total = sum(rank), median = median(Bench))

ggplot(data = female_bench_plot, aes(x = AgeClass, y = median)) +
  geom_point() +
  labs(y = "Median Ranking") +
  theme_minimal()
```



```{r}
# uncomment the two write statements to save the files
write_csv(openpl_male, "openpl_male.csv")
write_csv(openpl_female, "openpl_female.csv")

install.packages("openxlsx")
library(openxlsx)
write.xlsx(openpl_male, file = "openpl_male.xlsx")
```

## Familiarizing with Data

#### Lookiing at histograms for normality

```{r}
ggplot(data = openpl_female, aes(x = TotalKg)) +
  geom_histogram()
ggplot(data = openpl_female, aes(x = Squat)) +
  geom_histogram()
ggplot(data = openpl_female, aes(x = Deadlift)) +
  geom_histogram()
ggplot(data = openpl_female, aes(x = Bench)) +
  geom_histogram() +
  facet_wrap(~AgeClass)
## for women, bench seems slightly right skewed but 
## everything else looks close to normal

ggplot(data = openpl_male, aes(x = TotalKg)) +
  geom_histogram()
ggplot(data = openpl_male, aes(x = Squat)) +
  geom_histogram()
ggplot(data = openpl_male, aes(x = Deadlift)) +
  geom_histogram()
ggplot(data = openpl_male, aes(x = Bench)) +
  geom_histogram()
## same distributions for men, less skewed bench
```

#### Grouping by Age

```{r}
age_avgs_female <- openpl_female |> 
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  group_by(AgeClass) |>
  summarise(Squat = mean(Squat), Deadlift = mean(Deadlift), 
            Bench = mean(Bench), TotalKg = mean(TotalKg)) |>
  ungroup()

ggplot(data = age_avgs_female, aes(x = AgeClass, y = TotalKg)) +
  geom_point()
ggplot(data = age_avgs_female, aes(x = AgeClass, y = Squat)) +
  geom_point()
ggplot(data = age_avgs_female, aes(x = AgeClass, y = Deadlift)) +
  geom_point()
ggplot(data = age_avgs_female, aes(x = AgeClass, y = Bench)) +
  geom_point()
## for females, all exercises look like they peak at 24-34 or 35-39

age_avgs_male <- openpl_male |> 
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  group_by(AgeClass) |>
  summarise(Squat = mean(Squat), Deadlift = mean(Deadlift), 
            Bench = mean(Bench), TotalKg = mean(TotalKg)) |>
  ungroup()

ggplot(data = age_avgs_male, aes(x = AgeClass, y = TotalKg)) +
  geom_point()
ggplot(data = age_avgs_male, aes(x = AgeClass, y = Squat)) +
  geom_point()
ggplot(data = age_avgs_male, aes(x = AgeClass, y = Deadlift)) +
  geom_point()
ggplot(data = age_avgs_male, aes(x = AgeClass, y = Bench)) +
  geom_point()
## similar pattern with males, more consistent between exercises
## still peaks at 24-34 and 35-39
```

## Testing for normality

```{r}
female_bench <- openpl_female |> 
  mutate(Bench = as.numeric(Bench)) |>
  group_by(AgeClass) |>
  group_modify(~ if (nrow(.x) >= 50) sample_n(.x, 50) else .x) |>
  ungroup() |>
  select(Bench, Age)

ggplot(data = openpl_female, aes(x = Bench)) +
  geom_histogram() +
  facet_wrap(~AgeClass)
```

## Kruskal-Wallis Test

-   Used when

    -   population medians (null hypothesis is population medians are equal)

    -   the data is continuous or ordinal

    -   the assumptions (mainly normality) are violated for a one-way ANOVA

    -   observations are independent

    -   sample size is at least 5

-   How to run test

    -   form hypothesis

    -   rank data from smallest to largest from all groups

    -   find the sum of the ranks for each group

    -   use test in R to calculate the Kruskal-Wallis H statistic

    -   if p-value is less than 0.05, reject the null indicating as least one group differs slightly

    -   if null is rejected, use Dunn's test (pairwise comparison test) to find which groups specifically differ

## Running Test
Null hypothesis: there is no difference in medians ranking between age groups
Alternative hypothesis: there is at least one group that is different from the others
```{r}
female_bench_ranked <- openpl_female |> arrange(desc(Bench)) |>
  mutate(rank = row_number()) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  select(Name, Age, AgeClass, Bench, rank)

female_bench_ranked |>
  group_by(AgeClass)  |>
  summarise(n())

ggplot(data = female_bench_ranked, aes(x = AgeClass, y = Bench)) +
  geom_boxplot()

female_bench_plot <- female_bench_ranked |>
  group_by(AgeClass)  |>
  summarise(total = sum(rank), average = total/n())

ggplot(data = female_bench_plot, aes(x = AgeClass, y = average)) +
  geom_point() +
  scale_y_reverse() +
  theme_minimal()

female_bench_results <- kruskal.test(female_bench_ranked$rank ~ 
                                       female_bench_ranked$AgeClass)
female_bench_results
```

## Dunn's Test
```{r}
library(dunn.test)
library(FSA)
library(reshape2)
female_bench_dunn <- dunn.test(female_bench_ranked$rank, female_bench_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(female_bench_dunn$P.adjusted, female_bench_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `female_bench_dunn$P.adjusted`)

dunn_results_df$log_p <- -log10(dunn_results_df$p_value)
dunn_results_df$sig <- if_else(dunn_results_df$p_value <= 0.01,
                               1,0)
ggplot(dunn_results_df, aes(x = group1, y = group2, fill = sig)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal()

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = log_p)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal()

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = p_value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "red", high = "blue") +
  theme_minimal()

range(dunn_results_df$p_value)
```

https://rpkgs.datanovia.com/rstatix/reference/dunn_test.html
https://pubmed.ncbi.nlm.nih.gov/9413454/
This page uses data from the OpenPowerlifting project, https://www.openpowerlifting.org.
You may download a copy of the data at https://data.openpowerlifting.org.
