---
title: "Sample Powerlifting Module"
format: html
editor: visual
---
#### Load in appropriate packages and datasets:
```{r warning=FALSE, output=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
library(dunn.test)
library(FSA)
library(reshape2)

openpl_male <- read_csv("openpl_male.csv")

openpl_female <- read_csv("openpl_female.csv")
```

## Female Squat
#### Arrange data appropriately:
To run the Kruskal-Wallis test, the data needs to be arranged in a descending order (from highest weight lifted to lowest) and then ranked accordingly. Make a variable for rank using the mutate function.
```{r echo=TRUE, output = FALSE}
female_squat_ranked <- openpl_female |> arrange(desc(Squat)) |>
  mutate(rank = row_number()) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  select(Name, Age, AgeClass, Squat, rank)|>
  mutate(AgeClass = fct_recode(AgeClass,
                              "05-12" = "5-12"))
## recoding "5-12" as "05-12" so it appears in the correct order
```

#### Look at distribution of the data:
The boxplots can show us the general trend of the data by age class.
```{r echo=TRUE, eval=TRUE}
ggplot(data = female_squat_ranked, aes(x = AgeClass, y = Squat)) +
  geom_boxplot() +
  theme_minimal()
```

#### Form a hypothesis:
$H_0:$ there is no difference in the median of powerlifting weight between age groups

$H_a:$ at least one group differs from another

#### Run the test, confirm significant p-value:
When running the Kruskal-Wallis test for association we are looking to see if there is a p-value of less than 0.5. This would indicate there is at least one group that is significantly discernible from another.
```{r}
female_squat_results <- kruskal.test(female_squat_ranked$rank ~ 
                                       female_squat_ranked$AgeClass)
female_squat_results
```
We can confirm from this p-value that there is discernibility between groups so we want to run the Dunn's Test for multiple comparisons to see which groups are discernible from each other. 

#### Run pairwise test:
```{r output=FALSE}
female_squat_dunn <- dunn.test(female_squat_ranked$rank, female_squat_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(female_squat_dunn$P.adjusted, female_squat_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `female_squat_dunn$P.adjusted`)
```

#### Create a visualization to show log p-values:
Because the results from the test are difficult to visualize due to how many values there are, the best way to analyze the results is with a heatmap.
```{r}
dunn_results_log <- dunn_results_df |>
  mutate(log_p = -log10(p_value))

ggplot(dunn_results_log, aes(x = group1, y = group2, fill = log_p)) +
  geom_tile(colour = "black", linewidth = 0.5) +
  scale_fill_gradient(low = "forestgreen", high = "white") +
  theme_minimal()
```
The -log10 heatmap shows us how many zeros there are after the decimal point in the p-value. This indicates significance because the more zeros there are, the smaller the p-value is and the more significant the different is.

#### Create heatmap of significant p-values:
```{r}
dunn_results_df$sig <- if_else(dunn_results_df$p_value <= 0.05,1,0)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = sig)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_continuous(low = "grey100", high = "forestgreen") +
  theme_minimal()
```
This visualization 

## Female Deadlift
Fill in the neceessary work to run the same analysis for deadlifting.
#### Arrange data appropriately.
Use `mutate`, `select`, and `filter` functions to set up your data for the test.

#### Look at distribution of the data:
Create a boxplot by `AgeClass`.

#### Form a hypothesis:

#### Run the test, confirm significant p-value:

#### Run pairwise test:

#### Create a visualization to show log p-values:


#### Create heatmap of significant p-values:


## Male Powerlifting Analysis
Complete the following exercises using the male data set for `Bench`, `Squat`, and `Deadlift`

Exercise 1. Arrange the data appropriately to run the Kruskal-Wallis Test.

Exercise 2. Run the test and confirm whether is it appropriate to move forward.

Exercise 3, If appropriate, run the Dunn's Test to determine which classes are discernible.

Exercise 4. Create a visualization to show which classes are significantly discernible.

Exercise 5. Write an analysis of these results in your own words describing what your visualization tells us.
