---
title: "Sample Powerlifting Module"
format: html
editor: visual
---
## Female Squat
#### Load in appropriate packages and datasets:
```{r warning=FALSE, output=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
library(dunn.test)
library(FSA)
library(reshape2)

openpl_male <- read_csv("~/Desktop/SYE/github repos/stat450-spr2026-score/keelinjoyce/powerlifting//openpl_male.csv")

openpl_female <- read_csv("~/Desktop/SYE/github repos/stat450-spr2026-score/keelinjoyce/powerlifting//openpl_female.csv")
```

#### Arrange data appropriately:
```{r echo=TRUE, output = FALSE}
female_squat_ranked <- openpl_female |> arrange(desc(Squat)) |>
  mutate(rank = row_number()) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  select(Name, Age, AgeClass, Squat, rank)|>
  mutate(AgeClass = fct_recode(AgeClass,
                              "05-12" = "5-12"))
## recoding "5-12" as "05-12" so it appears in the correct order

female_squat_ranked |>
  group_by(AgeClass)  |>
  summarise(n())
```

#### Look at distribution of the data:
```{r echo=TRUE, eval=TRUE}
ggplot(data = female_squat_ranked, aes(x = AgeClass, y = Squat)) +
  geom_boxplot() +
  theme_minimal()
```

#### Run the test, confirm significant p-value:
```{r}
female_squat_results <- kruskal.test(female_squat_ranked$rank ~ 
                                       female_squat_ranked$AgeClass)
female_squat_results
```

#### Run pairwise test:
```{r output=FALSE}
female_squat_dunn <- dunn.test(female_squat_ranked$rank, female_squat_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(female_squat_dunn$P.adjusted, female_squat_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `female_squat_dunn$P.adjusted`)
```

#### Create a visualization to show log p-values:
```{r}
female_squat_dunn <- dunn.test(female_squat_ranked$rank, female_squat_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(female_squat_dunn$P.adjusted, female_squat_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `female_squat_dunn$P.adjusted`)

dunn_results_df$log_p <- -log10(dunn_results_df$p_value)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = p_value)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = "forestgreen", high = "grey100") +
  theme_minimal()
```

#### Create heatmap of significant p-values:
```{r}
dunn_results_df$sig <- if_else(dunn_results_df$p_value <= 0.05,1,0)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = sig)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_continuous(low = "grey100", high = "forestgreen") +
  theme_minimal()
```

## Female Deadlift
#### Arrange data appropriately:
```{r echo=TRUE, output = FALSE}
female_deadlift_ranked <- openpl_female |> arrange(desc(Deadlift)) |>
  mutate(rank = row_number()) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  select(Name, Age, AgeClass, Deadlift, rank)|>
  mutate(AgeClass = fct_recode(AgeClass,
                              "05-12" = "5-12"))
## recoding "5-12" as "05-12" so it appears in the correct order

female_deadlift_ranked |>
  group_by(AgeClass)  |>
  summarise(n())
```

#### Look at distribution of the data:
```{r echo=TRUE, eval=TRUE}
ggplot(data = female_deadlift_ranked, aes(x = AgeClass, y = Deadlift)) +
  geom_boxplot() +
  theme_minimal()
```

#### Run the test, confirm significant p-value:
```{r}
female_deadlift_results <- kruskal.test(female_deadlift_ranked$rank ~ 
                                       female_deadlift_ranked$AgeClass)
female_deadlift_results
```

#### Run pairwise test:
```{r output=FALSE}
female_deadlift_dunn <- dunn.test(female_deadlift_ranked$rank, female_deadlift_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(female_deadlift_dunn$P.adjusted, female_deadlift_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `female_deadlift_dunn$P.adjusted`)
```

#### Create a visualization to show log p-values:
```{r}
female_deadlift_dunn <- dunn.test(female_deadlift_ranked$rank, female_deadlift_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(female_deadlift_dunn$P.adjusted, female_deadlift_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `female_deadlift_dunn$P.adjusted`)

dunn_results_df$log_p <- -log10(dunn_results_df$p_value)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = p_value)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = "forestgreen", high = "grey100") +
  theme_minimal()
```

#### Create heatmap of significant p-values:
```{r}
dunn_results_df$sig <- if_else(dunn_results_df$p_value <= 0.05,1,0)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = sig)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_continuous(low = "grey100", high = "forestgreen") +
  theme_minimal()
```

## Male Squat
#### Arrange data appropriately:
```{r echo=TRUE, output = FALSE}
male_squat_ranked <- openpl_male |> arrange(desc(Squat)) |>
  mutate(rank = row_number()) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  select(Name, Age, AgeClass, Squat, rank)|>
  mutate(AgeClass = fct_recode(AgeClass,
                              "05-12" = "5-12"))
## recoding "5-12" as "05-12" so it appears in the correct order

male_squat_ranked |>
  group_by(AgeClass)  |>
  summarise(n())
```

#### Look at distribution of the data:
```{r echo=TRUE, eval=TRUE}
ggplot(data = male_squat_ranked, aes(x = AgeClass, y = Squat)) +
  geom_boxplot() +
  theme_minimal()
```

#### Run the test, confirm significant p-value:
```{r}
male_squat_results <- kruskal.test(male_squat_ranked$rank ~ 
                                       male_squat_ranked$AgeClass)
male_squat_results
```

#### Run pairwise test:
```{r output=FALSE}
male_squat_dunn <- dunn.test(male_squat_ranked$rank, male_squat_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(male_squat_dunn$P.adjusted, male_squat_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `male_squat_dunn$P.adjusted`)
```

#### Create a visualization to show log p-values:
```{r}
male_squat_dunn <- dunn.test(male_squat_ranked$rank, male_squat_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(male_squat_dunn$P.adjusted, male_squat_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `male_squat_dunn$P.adjusted`)

dunn_results_df$log_p <- -log10(dunn_results_df$p_value)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = p_value)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = "forestgreen", high = "grey100") +
  theme_minimal()
```

#### Create heatmap of significant p-values:
```{r}
dunn_results_df$sig <- if_else(dunn_results_df$p_value <= 0.05,1,0)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = sig)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_continuous(low = "grey100", high = "forestgreen") +
  theme_minimal()
```

## Male Deadlift
#### Arrange data appropriately:
```{r echo=TRUE, output = FALSE}
male_deadlift_ranked <- openpl_male |> arrange(desc(Deadlift)) |>
  mutate(rank = row_number()) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  select(Name, Age, AgeClass, Deadlift, rank)|>
  mutate(AgeClass = fct_recode(AgeClass,
                              "05-12" = "5-12"))
## recoding "5-12" as "05-12" so it appears in the correct order

male_deadlift_ranked |>
  group_by(AgeClass)  |>
  summarise(n())
```

#### Look at distribution of the data:
```{r echo=TRUE, eval=TRUE}
ggplot(data = male_deadlift_ranked, aes(x = AgeClass, y = Deadlift)) +
  geom_boxplot() +
  theme_minimal()
```

#### Run the test, confirm significant p-value:
```{r}
male_deadlift_results <- kruskal.test(male_deadlift_ranked$rank ~ 
                                       male_deadlift_ranked$AgeClass)
male_deadlift_results
```

#### Run pairwise test:
```{r output=FALSE}
male_deadlift_dunn <- dunn.test(male_deadlift_ranked$rank, male_deadlift_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(male_deadlift_dunn$P.adjusted, male_deadlift_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `male_deadlift_dunn$P.adjusted`)
```

#### Create a visualization to show log p-values:
```{r}
male_deadlift_dunn <- dunn.test(male_deadlift_ranked$rank, male_deadlift_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(male_deadlift_dunn$P.adjusted, male_deadlift_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `male_deadlift_dunn$P.adjusted`)

dunn_results_df$log_p <- -log10(dunn_results_df$p_value)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = p_value)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = "forestgreen", high = "grey100") +
  theme_minimal()
```

#### Create heatmap of significant p-values:
```{r}
dunn_results_df$sig <- if_else(dunn_results_df$p_value <= 0.05,1,0)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = sig)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_continuous(low = "grey100", high = "forestgreen") +
  theme_minimal()
```

## Male Bench
#### Arrange data appropriately:
```{r echo=TRUE, output = FALSE}
male_bench_ranked <- openpl_male |> arrange(desc(Bench)) |>
  mutate(rank = row_number()) |>
  filter(!is.na(AgeClass)) |>
  mutate(AgeClass = fct_relevel(AgeClass, "5-12")) |>
  select(Name, Age, AgeClass, Bench, rank)|>
  mutate(AgeClass = fct_recode(AgeClass,
                              "05-12" = "5-12"))
## recoding "5-12" as "05-12" so it appears in the correct order

male_bench_ranked |>
  group_by(AgeClass)  |>
  summarise(n())
```

#### Look at distribution of the data:
```{r echo=TRUE, eval=TRUE}
ggplot(data = male_bench_ranked, aes(x = AgeClass, y = Bench)) +
  geom_boxplot() +
  theme_minimal()
```

#### Run the test, confirm significant p-value:
```{r}
male_bench_results <- kruskal.test(male_bench_ranked$rank ~ 
                                       male_bench_ranked$AgeClass)
male_bench_results
```

#### Run pairwise test:
```{r output=FALSE}
male_bench_dunn <- dunn.test(male_bench_ranked$rank, male_bench_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(male_bench_dunn$P.adjusted, male_bench_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `male_bench_dunn$P.adjusted`)
```

#### Create a visualization to show log p-values:
```{r}
male_bench_dunn <- dunn.test(male_bench_ranked$rank, male_bench_ranked$AgeClass, method = "bh")
dunn_results_df <- as.data.frame(male_bench_dunn$P.adjusted, male_bench_dunn$comparisons) |>
  tibble::rownames_to_column("comparison") |>
  separate(comparison, into = c("group1", "group2"), sep = " - ") |>
  rename("p_value" = `male_bench_dunn$P.adjusted`)

dunn_results_df$log_p <- -log10(dunn_results_df$p_value)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = p_value)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = "forestgreen", high = "grey100") +
  theme_minimal()
```

#### Create heatmap of significant p-values:
```{r}
dunn_results_df$sig <- if_else(dunn_results_df$p_value <= 0.05,1,0)

ggplot(dunn_results_df, aes(x = group1, y = group2, fill = sig)) +
  geom_tile(color = "black", linewidth = 0.5) +
  scale_fill_continuous(low = "grey100", high = "forestgreen") +
  theme_minimal()
```

