---
title: title: "Kung Fu - Data manipulation, data visualization and Hypothesis Testing"
author: Elena Cortina Pino
date: February 19, 2026
format: html
---

### Load libraries

```{r}
#| message: false
#| warning: false

# Loading the libraries in:
library(tidyverse)
library(ggplot2)

```

### Know Thy Data

```{r}
kungfu<- read.csv("kungfu_data.csv")
head(kungfu, 25)
```

Because of the way the raw data was presented in the result book, the parsing process helped us, but only so much! We have a lot of information in the result book data about the athletes and the organization of the competition that we can use for the data analysis process, but it is not in a useful format in the current way it's presented after the parsing process.

We'll need to do some significant `stringr` work to separate the information in column labeled `Form` into:\
1. Athlete sex\
2. Kung fu form\
3. Group

```{r}
#| message: false
#| warning: false


# this took forever to figure out 

kungfu_strings <- kungfu |> select(-Page) |> filter(Form != "Bare-hand Duilian(2 Persons)", Form != "Weapon Duilian(2 Persons)", Form != "Other Duilian") |> # this line gets rid of the form "duilian" because that's a group event
  separate_wider_regex(Form, c(sex = "^.", Form = ".*")) # this made the Sex column with the first letter of the words "Men's" and "Women's"
```

```{r}
#| message: false
#| warning: false

# this code gets rid of the end of "en's and "omen's"
getting_rid_of_up_to_form <- kungfu_strings |> separate_wider_regex(Form,
  patterns = c(
    delete = "^[^\\s]+",
    "\\s+",
    rest = ".*"))

# we're getting there 
```



```{r}
#| message: false
#| warning: false

# this code creates the form column by selecting everything up to the capital G for "Group" and by doing that creates the column for Group

selecting_the_form <- getting_rid_of_up_to_form |> select(-delete) |>  separate_wider_regex(rest,
  patterns = c(
    Form = ".*?(?=G)", 
    Group = ".*"))
```
Now the situation is manageable. The remaining step is carpentry that repairs the missing `Group` category we lost during the parsing process of the result book.\

```{r}
# this line gives me a csv file that I will add letters to the groups that are missing the letter by hand
write.csv(selecting_the_form, "write_group_letter.csv")
```

Add the new csv back into here
```{r}
lets_go <- read.csv(groups_added_kungfu.csv)
```




