---
title: "kungfu_solution"
desciption: Solution file to the activity proposed in the Main Kung Fu Complete Linkage Clustering Heatmap Module
format: html
---

::: {.callout-note appearance="simple}
This is the solution file to the activity proposed in ["2025 10th World Kung Fu Championship - Complete Linkage Hierarchical Clustering" module](insert link here for the other module), the data to be used is [data for the 8th edition of the WKFC (2017)](year8kungfudata.csv)
:::

### Activity Solution

_The following solution is simply one of the many ways (as explained in the module) that exist to successfully and efficiently construct a complete linkage clustered heatmap for competitive Kung Fu data._


```{r}
#| label: libraries
#| message: false
#| echo: false
#| warning: false

# Loading the libraries in:

library(tidyverse)
library(ggplot2)
library(here)
library(VIM)
library(gplots)
library(pheatmap)
library("RColorBrewer")
library(grid)
```

```{r}
#| label: datareading
#| warning: false
#| echo: false
#| eval: true

# read data which comes from many many iterations of dplyr code
year8data <- read.csv("year8kungfudata.csv") |> select(-X)

# selecting OUT non-numeric columns in the data frame
mat <- as.matrix(year8data[,-1])

# chooses which column from the dataset to use for row names
rownames(mat) <- as.character(year8data$Form_new)

# setting color palette to blue, white, and red from WKFC logo colors
col<- colorRampPalette(c("navy", "white", "orange"))(256)
```

```{r}
#| label: heatmap
#| warning: false
#| eval: false


# using pheatmap() to create the clustered heatmap
heatmap <- pheatmap(
  mat,
  scale = "row", # values centered and scaled in row direction 
  clustering_distance_rows = 'euclidean',
  #clustering_distance_cols = 'euclidean',
  clustering_method = "complete",
  cluster_rows = T,
  cluster_cols = F,
  show_colnames = TRUE, 
  show_rownames = TRUE,
  display_numbers = T,
  cellwidth = NA,
  cellheight = NA, 
  angle_col = 0,
  number_color = "#303030",
  border_color = "#303030", 
  # main = "Complete Linkage Clustered Heatmap with Dendrograms",
  cutree_rows = 5, # number of major clusters of forms. 6 is too many
  color = col,
  legend = T
  )

```

Really interesting insights from this plot! Specifically if we zoom into the smaller clusters:\
\
1. Yang Style Taijiquan (杨式太极拳).\
2. Zui Jian (Drunken Sword) (醉贱).\
    Only had 1 participating group with 7 athletes competing.\
3.1 Chen Style Taijiquan (陈式太极拳).\
3.2 Other Traditional Single Weapon\
3.3 Jian (Straight Sword) (剑).
