---
title: title: "Kung Fu"
subtitle: Complete Linkage Hierarchical Clustering
author: Elena Cortina Pino
date: February 19, 2026
format: html
---

### Load libraries

```{r}
#| message: false
#| warning: false

# Loading the libraries in:
library(tidyverse)
library(ggplot2)
library(gplots)
library(pheatmap)
library(naniar)
library(mice)
library(VIM)
library(gplots)
library(pheatmap)
library("RColorBrewer")
library(grid)
```


### Know Thy Data

```{r}
kungfu<- read.csv("kungfu_data_dirty.csv")
head(kungfu, 25)
```

Because of the way the raw data was presented in the result book, the parsing process helped us, but only so much! We have a lot of information in the result book data about the athletes and the organization of the competition that we can use for the data analysis process, but it is not in a useful format in the current way it's presented after the parsing process.

We'll need to do some significant `stringr` work to separate the information in column labeled `Form` into:\
1. Athlete sex\
2. Kung fu form\
3. Group

```{r}
#| message: false
#| warning: false


# this took forever to figure out 

kungfu_strings <- kungfu |> select(-Page) |> filter(Form != "Bare-hand Duilian(2 Persons)", Form != "Weapon Duilian(2 Persons)", Form != "Other Duilian") |> # this line gets rid of the form "duilian" because that's a group event
  separate_wider_regex(Form, c(sex = "^.", Form = ".*")) # this made the Sex column with the first letter of the words "Men's" and "Women's"
```

```{r}
#| message: false
#| warning: false

# this code gets rid of the end of "en's and "omen's"
getting_rid_of_up_to_form <- kungfu_strings |> separate_wider_regex(Form,
  patterns = c(
    delete = "^[^\\s]+",
    "\\s+",
    rest = ".*"))

# we're getting there 
```

```{r}
#| message: false
#| warning: false

# this code creates the form column by selecting everything up to the capital G for "Group" and by doing that creates the column for Group

selecting_the_form <- getting_rid_of_up_to_form |> select(-delete) |>  separate_wider_regex(rest,
  patterns = c(
    Form = ".*?(?=G)", 
    Group = ".*"))
```

Now the situation is manageable. The remaining step is carpentry that repairs the missing `Group` category we lost during the parsing process of the result book.\

Add the new final clean everything correct and checked thrice csv back into here: the data frame is called `kf`

```{r}
kf <- read.csv("kungfu.csv")
```

Now we can start working with the data set to decide what to delete and what to continue developing.

Things we can do:

1.  get rid of the entries for athletes who defaulted.

-   Defaulting in the competition means they either got disqualified or had to step away from the competition voluntarily. Both outcomes resulted in them not getting scored.

2.  get summary statistics for both gender groups.

3.  get summary statistics for each (of the 6) age groups.

4.  get summary statistics for each kung fu form in the competition.

5.  get summary statistics for each country participating.

### Part by part

```{r}
# the number of total athletes before removing defaulted athletes is 2129
sample_kungfu <- kf|> filter(Rank != "Default") |> drop_na(Score) |> select(-X, -Rank, -Remark)
# the number of athletes after removing defaulting athletes is 1989
# difference is 140 athletes.
```


Writing sample_kungfu to make it usable to fix the form names

```{r}
#| eval: false
#| echo: false


# write.csv(sample_kungfu, "sample_kungfu.csv")
```


```{r}
usable_sample_kungfu <- read.csv("sample_kungfu.csv") |> mutate(recoded_forms = fct_recode(Form,
  # Shanzi variations
  "Single weapon routines Shanzi (Fan)" = "Single weapon routines Shanzi (Fan)",
  "Single weapon routines Shanzi (Fan)" = "Single weapon routines shanzi",
  
  # Pudao variations
  "Single weapon routines Pudao" = "Single weapon routines Pudao",
  "Single weapon routines Pudao" = "Single weapon routines Pudao/Dadao",
  "Single weapon routines Pudao" = "Single weapon routines Pudao/Guandao (Kwan Dao)",
  "Single weapon routines Pudao" = "Single weapon routines Guandao /Pudao",
  "Single weapon routines Pudao" = "Single weapon routines Guandao (Kwan Dao)",
  
  # Weapon other variations
  "Other weapon routines" = "weapon Other routines",
  "Other weapon routines" = "weapon routines Other routines",
  
  # Bare hand other variations
  "Other bare hand routines" = "bare hand routines Other routines",
  "Other bare hand routines" = "Other bare hand routines ",
  "Other bare hand routines" = "bare hand routines Other routines",
  
  # Taiji weapon variations
  "Single weapon routines Other Taiji weapon" = "Single weapon routines Other taiji weapon",
  "Single weapon routines Other Taiji weapon" = "Single weapon routines Other Taiji weapon"
)) |> group_by(Group, recoded_forms) |> summarise(mean_score = mean(Score), .groups = "drop")|> pivot_wider(names_from = Group, values_from = mean_score)

```


Imputation: Chosen method KNN

```{r}
kf10_imp_knn <- kNN(usable_sample_kungfu, k = 5) |> select(-A_imp,-B_imp,-C_imp,-D_imp,-E_imp,-F_imp, -recoded_forms_imp)  # You can adjust 'k' as needed
```


```{r}

vis_miss(usable_sample_kungfu)
```

imputing through MICE method:

```{r}
# Set the seed for reproducibility
set.seed(12345)

# Perform Multiple Imputation
imputed_mice <- mice(usable_sample_kungfu, m=5, method='pmm', print=FALSE)
```

\
Number of Imputations (m=5): The 'm' argument specifies how many complete datasets you wish to generate, each with missing values filled in. By setting m to 5, the function will create five versions of your dataset, each with missing values imputed differently. This multiplicity captures the uncertainty inherent in the imputation process. \
\
Imputation Method (method = 'pmm'): The 'method' argument dictates the statistical technique mice will use to predict missing values. PMM means Predictive Mean Matching and it's a non-parametric approach particularly suited for continuous data. PMM operates by finding observed values with similar predictive characteristics to the missing entries. The missing values are then imputed u, thus preserving the distribution and variance of the original data more effectively than simpler methods, such as mean imputation.\

```{r}
write.csv(kf10_imp, "mat_10kf.csv")
```


#### Hierarchical agglomerative clustering


[reference link for how to do the heatmap](https://www.datanovia.com/en/lessons/heatmap-in-r-static-and-interactive-visualization/)


```{r}

# this is the code from chatgpt that gets the means for the matrix
# below the code for the matrix is also from chat
hehe_kungfu <- sample_kungfu |> group_by(Group, Form) |> summarise(mean_score = mean(Score), .groups = "drop")|> pivot_wider(names_from = Group, values_from = mean_score)

# then i wrote this as a new file and changed some stuff... 
# we will never know what i changed because i also dont know
# but i changed group names by hand , wont do it again

# write.csv(hehe_kungfu, "10kf.csv")
```


```{r}
kf10 <- read.csv("10kf.csv")
```


```{r}
mat <- as.matrix(kf10[,-1])
rownames(mat) <- kf10$Form
mat[is.na(mat)] <- 8.50
# replace na's with means either from groups or forms

print(mat)

# now we will make the heatmap (accept a mess due to zeroes)
# or code zeroes to be black or neon, something that doesn't relate to 
# the rest of the heatmap's color spectrum

```

Use the complete function from tidyr to fill in missing factor combinations in your data.frame with NA.

Then you can use the na.value parameter in the color gradient to set the color.



**Revise this  source for clustering**:

https://rlbarter.github.io/superheat/clustering.html



```{r}
#| label: heatmap
#| warning: false


heatmap <- pheatmap(
  mat,
  scale = "row", # character indicating if the values should be centered and scaled in either the row direction or the column direction, or none.
  clustering_distance_rows = 'canberra',
  clustering_distance_cols = 'canberra',
  clustering_method = "complete",
  cluster_rows = T,
  cluster_cols = T,
  show_colnames = TRUE, 
  show_rownames = TRUE,
  fontsize_row = 7,
  fontsize_col = 7, 
  angle_col = 0,
  main = "Heatmap with dendrogram",
  cutree_rows = 5,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100),
  legend = T)


#clusters<- heatmap$tree_row

```




```{r}
#| warning: false
#| eval: false
#| message: false
#| label: references

# reference code
heatmap(matrix_kf, scale = "column")
# custom color palette sample code
col<- colorRampPalette(c("azure", "white", "#e00707"))(256)
library("RColorBrewer")




heat_plot <- pheatmap(data, 
                      col = brewer.pal(10, 'RdYlGn'), # choose a colour scale for your data
                      cluster_rows = T, cluster_cols = T, # set to FALSE if you want to remove the dendograms
                      clustering_distance_cols = 'euclidean',
                      clustering_distance_rows = 'euclidean',
                      clustering_method = 'ward.D',
                      annotation_row = gene_functions_df, # row (gene) annotations
                      annotation_col = ann_df, # column (sample) annotations
                      annotation_colors = ann_colors, # colours for your annotations
                      annotation_names_row = F, 
                      annotation_names_col = F,
                      fontsize_row = 10,          # row label font size
                      fontsize_col = 7,          # column label font size 
                      angle_col = 45, # sample names at an angle
                      legend_breaks = c(-2, 0, 2), # legend customisation
                      legend_labels = c("Low", "Medium", "High"), # legend customisation
                      show_colnames = T, show_rownames = F, # displaying column and row names
                      main = "Super heatmap with annotations") # a title for our heatmap
```

```{r}
kungfu_7 <- read.csv("kungfu_7.csv")
kungfu_8 <- read.csv("kungfu_8.csv")
kungfu_9 <- read.csv("kungfu_9.csv")

```

```{r}
working_eight <- kungfu_8 |> group_by(Form) |> summarise(haha = n()) |> select(Form)


write.csv(working_eight, "8 events.csv")
```

```{r}
working_nine <- kungfu_9 |> group_by(Form) |> summarise(haha = n()) |> select(Form)
write.csv(working_nine, "9 events.csv")

```

```{r}
working_ten <- kf |> group_by(Form) |> summarise(haha = n()) |> select(Form)
write.csv(working_ten, "10 events.csv")
```



```{r}
#| label: another_situation
# GETTING RID OF YEAR 9 AND EXPLAIN WHY

activity_year9 <- read.csv("activity_kungfu_year9.csv") |> group_by(Group, Form) |> summarise(mean_score = mean(Score), .groups = "drop")|> pivot_wider(names_from = Group, values_from = mean_score)
```



```{r}
year8goat <- read.csv("year8forsure.csv") |> drop_na(Score) |> mutate(Form_new = fct_recode(Form,
  "Imitation Eagle Style" = "Imitation Eagle Style",
  "Imitation Eagle Style" = "Imitation Eagle Styles",
  
  "Other Taiji Weapon" = "Other Style Taiji Weapon",
  "Other Taiji Weapon" = "Other Taiji Weapon",
  
  "Other Traditional Flexible/Soft Weapon" = "Other Traditional Flexible/Soft Weapon",
  "Other Traditional Flexible/Soft Weapon" = "Other Traditional Flexible/Soft Weapon F",
  
  "Yang Style Taiji Jian" = "Yang StyleTaiji Jian",
  "Yang Style Taiji Jian" = "Yang Style Taiji Jian",
  
  "Other Traditional Single Weapon" = "Other Traditional Single Weapon",
  "Other Traditional Single Weapon" = "Other Traditional Single Weapons",
  "Other Traditional Single Weapon" = "Other Traditional Weapon",
  "Other Traditional Single Weapon" = "Other Weapon")) |> group_by(Group, Form_new) |> summarise(mean_score = mean(Score), .groups = "drop")|> pivot_wider(names_from = Group, values_from = mean_score) |> select(-A) |> select(-`F`) |>  select(-B)
```




```{r}

#| label: another_situation

year8_imp_knn <- kNN(year8goat, k = 5) |> select(-C_imp,-D_imp,-E_imp, -`Form_new_imp`)

write.csv(year8_imp_knn, "year8kungfudata.csv")
```




