---
title: "Doughnut Run: Time Adjustments & Performance"
format: html
---

In some races (especially fun runs), runners may stop for bonus challenges—like eating doughnuts. This worksheet explores:

**How many doughnuts did runners eat, and how is that related to race time?**

You will work with three datasets:

- `doughnut2015.csv`: adjusted race results
- `doughnut2015unadj.csv`: unadjusted race results
- `doughnuttime.csv`: the “bonus time thresholds” for each doughnut

You’ll practice realistic data analyst tasks:
1. cleaning messy time strings
2. converting time into seconds for analysis
3. joining datasets by an ID
4. turning thresholds into a derived variable (donuts eaten)
5. visualizing patterns and fitting a simple model

# 0. Load the Following Packages

```{r}
#| message: false
library(tidyverse)
library(lubridate)
```


# 1. Load in the CSV Files
Load the three datasets from your data folder.

```{r}
doughnut2015     <- read_csv("Tim_Kashkarov/doughnut2015.csv")
doughnut2015unadj<- read_csv("Tim_Kashkarov/doughnut2015unadj.csv")
doughnuttime     <- read_csv("Tim_Kashkarov/doughnuttime.csv")
```


# 2. Fix the Time Column Format
In these datasets, the Time column may be inconsistent. Some values look like mm:ss:00 but should really be interpreted as 00:mm:ss.

Your goal: make sure every time is in HH:MM:SS format.

a) Write a function fix_time() that:
1. turns mm:ss:00 into 00:mm:ss
2. pads h:m:s into hh:mm:ss
3. returns NA if the format is not usable

```{r}
fix_time <- function(x) {
  x <- as.character(x)
  parts <- strsplit(x, ":", fixed = TRUE)

  vapply(parts, function(p) {
    if (length(p) != 3) return(NA_character_)

    # if third element is "00", treat as mm:ss:00 -> 00:mm:ss
    if (p[3] == "00") {
      sprintf("00:%02d:%02d", as.integer(p[1]), as.integer(p[2]))
    } else {
      sprintf("%02d:%02d:%02d", as.integer(p[1]), as.integer(p[2]), as.integer(p[3]))
    }
  }, character(1))
}
```

b) Apply the function to both race result datasets

```{r}
doughnut2015 <- doughnut2015 |>
  mutate(Time = fix_time(Time))

doughnut2015unadj <- doughnut2015unadj |>
  mutate(Time = fix_time(Time))
```


# 3. Convert HH:MM:SS to Total Seconds
For analysis, it is easier to compare times as numbers (seconds).

a) Create a new column time_sec in each race dataset
```{r}
doughnut2015 <- doughnut2015 |>
  mutate(time_sec = as.numeric(as.difftime(Time, format = "%H:%M:%S", units = "secs")))

doughnut2015unadj <- doughnut2015unadj |>
  mutate(time_sec = as.numeric(as.difftime(Time, format = "%H:%M:%S", units = "secs")))
```


# 4.
