---
title: "answerkey"
format: html
---

Parity analysis of NHL and KHL.

In sports analytics, a common question is parity: how evenly matched are teams in a league.

This worksheet explores:
*Are teams more evenly matched in the NHL or the KHL (2024–25)?*

You will work with two datasets:

- "nhl_20245.csv":  NHL team summary stats for 2024–25

- "khl_20245.csv": KHL team summary stats for 2024–25

You’ll practice realistic data analyst tasks:

1. creating comparable variables across datasets
2. visualizing distributions to compare leagues
3. computing spread (parity) metrics like IQR and P90–P10
4. using a randomization (permutation) test to evaluate differences

# 0. Load the Following Packages

```{r}
#| message: false
library(tidyverse)
library(lubridate)
```

# 1. Load in the CSV Files

```{r}
nhl_20245 <- read_csv("Tim_Kashkarov/nhl_khl/nhl_20245.csv")
khl_20245 <- read_csv("Tim_Kashkarov/nhl_khl/khl_20245.csv")
```

# 2. Create Comparable Variables

```{r}
nhl_20245 <- nhl_20245 |>
  transmute(
    league = "NHL",
    team = Team,
    WinPerc = Wins / GamesPlayed,
    GF_pg = GF / GamesPlayed,
    GA_pg = GA / GamesPlayed
  )

khl_20245 <- khl_20245 |>
  transmute(
    league = "KHL",
    team = Team,
    WinPerc = Wins / GamesPlayed,
    GF_pg = GF / GamesPlayed,
    GA_pg = GA / GamesPlayed
  )

df <- bind_rows(nhl_20245, khl_20245)
```

# 3. Randomization: What Are We Trying to Prove?

When we compare leagues, we will see some difference no matter what. The real question is:

*Is the difference we see big enough that it isn’t random?*

That’s what a randomization test helps us decide.

We will:

1. Measure parity in the real data.
2. Calculate the real difference: KHL - NHL.
3. Shuffle the league labels (NHL/KHL) many times.
4. After each shuffle, recompute the difference.
5. See how often shuffled differences are as large as the real one.

If the real difference is rare under shuffling -> evidence that the leagues are truly different.

# 4. Visualize Win% Distributions (First Look at Parity)

Parity means win percentages are more “bunched together”.

a) Histogram and density plot
```{r}
ggplot(df, aes(x = WinPerc, fill = league)) +
  geom_histogram(aes(y = after_stat(density)), bins = 12, alpha = 0.4, position = "identity") +
  geom_density(alpha = 0.2) +
  facet_wrap(~league, ncol = 1) +
  labs(
    title = "Win% Distribution (2024/25)",
    x = "Win Percentage (W / GP)",
    y = "Density"
  ) +
  theme_minimal()
```

b) Boxplot
```{r}
ggplot(df, aes(x = league, y = WinPerc)) +
  geom_boxplot() +
  labs(
    title = "Win% Spread: NHL vs KHL (2024/25)",
    x = "",
    y = "Win Percentage (W/GP)"
  ) +
  theme_minimal()
```

# 5. Choose a Parity Metric: Start with IQR

A parity metric is just a number that describes “spread".

We’ll start with:

IQR (interquartile range) = Q75 − Q25
- spread of the middle 50% of teams
- smaller IQR = more parity

a) Observed IQR difference
$$obs = IQR(KHL) - IQR(NHL)$$
```{r}
obs_iqr <- with(df, IQR(WinPerc[league=="KHL"]) - IQR(WinPerc[league=="NHL"]))
obs_iqr
```

b) Run the randomization test (shuffle labels)

```{r}
set.seed(1)

B <- 10000
diffs_iqr <- numeric(B)

for(i in 1:B){
  shuffled <- df |>
    mutate(league_shuf = sample(league))
  
  diffs_iqr[i] <- with(shuffled,
    IQR(WinPerc[league_shuf=="KHL"]) - IQR(WinPerc[league_shuf=="NHL"])
  )
}

p_val_iqr <- mean(diffs_iqr >= obs_iqr)
p_val_iqr
```

How to interpret the p-value:

small p-value (< 0.05) -> real difference is unusual under shuffling
bigger p-value -> difference could easily happen by chance

# 6. What Could Be “Wrong” With Shuffling?

Shuffling assumes something called exchangeability:

*If there is no real league difference, then it should not matter which teams are labeled NHL vs KHL.*

But in real sports, leagues can differ in ways that make this imperfect, such as:

1. different number of teams
2. different schedule format
3. different overall level of competition
4. real leagues have a balance of W/L, fake leagues have not

So shuffling is a useful “null model,” but it’s still an assumption.

That’s why we’re going to introduce a second idea: Bradley–Terry Concept.



# 7. Bradley–Terry Concept: Converting Strength into Win Probability

Bradley–Terry is a model for pairwise matchups:
$$P(i\,beats\,j) = \frac{s_{i}}{s_{i} + s_{j}}$$
- If two teams have equal strength -> probability = 0.5
- If one team is stronger -> probability > 0.5

So parity can be described as:

- strengths close together -> probabilities near 0.5 -> high parity
- strengths spread out -> probabilities extreme -> low parity

*Toy example*
```{r}
strength <- c(A = 0.70, B = 0.55, C = 0.40)

c(
  P_A_beats_B = strength["A"] / (strength["A"] + strength["B"]),
  P_A_beats_C = strength["A"] / (strength["A"] + strength["C"]),
  P_B_beats_C = strength["B"] / (strength["B"] + strength["C"])
)
```

In real leagues every win is someone else’s loss. When we shuffle labels, we create “fake leagues” that may not preserve that natural win/loss balance structure within each league, so the shuffled groups can be unrealistic.

# 8. A BT-Inspired Strength Score (Using WinPerc)

We do not have head-to-head game results here, so we will not fit a full Bradley–Terry model.

Instead, we build a BT-inspired score using WinPerc:
$$BT_i = \frac{1}{n-1}\sum_{j \ne i}\frac{w_i}{w_i + w_j}$$
This gives each team an “average pairwise win probability” against all other teams.

a) Write a function
```{r}
BT_scores <- function(w) {
  n <- length(w)
  bt <- numeric(n)

  for(i in 1:n){
    bt[i] <- mean(w[i] / (w[i] + w[-i]))
  }
  bt
}
```

b) Compute BT scores and measure their spread
```{r}
df_bt <- df |>
  group_by(league) |>
  mutate(BT = BT_scores(WinPerc)) |>
  ungroup()

df_bt |>
  group_by(league) |>
  summarise(
    BT_IQR = IQR(BT),
    BT_P90_P10 = quantile(BT, 0.90) - quantile(BT, 0.10),
    .groups = "drop"
  )
```

# 9. Randomization Test Again (Now Using BT_IQR)

Now we repeat the randomization, but this time our parity metric is BT_IQR.

a) Observed difference
```{r}
obs_bt_iqr <- with(df_bt, IQR(BT[league=="KHL"]) - IQR(BT[league=="NHL"]))
obs_bt_iqr
```

b) Shuffle labels and recompute
```{r}
set.seed(1)

B <- 5000
diffs_bt <- numeric(B)

for(i in 1:B){
  shuffled <- df_bt |>
    mutate(league_shuf = sample(league))

  diffs_bt[i] <- with(shuffled,
    IQR(BT[league_shuf=="KHL"]) - IQR(BT[league_shuf=="NHL"])
  )
}

p_val_bt <- mean(diffs_bt >= obs_bt_iqr)
p_val_bt
```

# 10. Final Exercise: Use Multiple Parity Metrics

Now you will measure parity using multiple metrics (not just IQR).

Compute:

1. IQR
3. P90 − P10

```{r}
parity_metrics <- df |>
  group_by(league) |>
  summarise(
    IQR = IQR(WinPerc),
    P90_P10 = quantile(WinPerc, 0.90) - quantile(WinPerc, 0.10),
    .groups = "drop"
  )

parity_metrics
```

# Your Turn 

a) For each metric (IQR, P90–P10), which league looks like it has more parity?

b) Pick P90–P10 and run a permutation test:

- compute the observed difference (KHL − NHL)
- shuffle labels many times
- compute the p-value

c) Write 2–3 sentences interpreting your result.

- Why do we shuffle the league labels instead of shuffling WinPerc values?
- What does the permutation test help you conclude?
- What is the key idea behind the Bradley–Terry probability formula?


