---
title: "draft"
format: html
---

```{r}
#| message: false
library(tidyverse)
library(lubridate)
```

```{r}
nhl_20245 <- read_csv("Tim_Kashkarov/nhl_khl/nhl_20245.csv")
khl_20245 <- read_csv("Tim_Kashkarov/nhl_khl/khl_20245.csv")
```

```{r}
nhl_20245 <- nhl_20245 |> transmute(league = "NHL", team = Team, WinPerc = Wins / GamesPlayed, GF_pg = GF / GamesPlayed, GA_pg = GA / GamesPlayed)
khl_20245 <- khl_20245 |> transmute(league = "KHL", team = Team, WinPerc = Wins / GamesPlayed, GF_pg = GF / GamesPlayed, GA_pg = GA / GamesPlayed)
df <- bind_rows(nhl_20245, khl_20245)
```

```{r}
ggplot(df, aes(x = WinPerc, fill = league)) +
  geom_histogram(aes(y = after_stat(density)), bins = 12, alpha = 0.4, position = "identity") +
  geom_density(alpha = 0.2) +
  facet_wrap(~league, ncol = 1) +
  labs(title = "Win Distribution (2024/25)", x = "Win (W / GP)", y = "Density") +
  theme_minimal()

ggplot(df, aes(x = league, y = WinPerc)) +
  geom_boxplot() +
  labs(title = "Win spread: NHL vs KHL (2024/25)", x = "", y = "Win (W/GP)") +
  theme_minimal()
```

```{r}
spread <- df |>
  group_by(league) |>
  summarise(
    teams = n(),
    min = min(WinPerc),
    max = max(WinPerc),
    range = max - min,
    IQR = IQR(WinPerc),
    P90_P10 = quantile(WinPerc, 0.90) - quantile(WinPerc, 0.10),
    .groups = "drop")
spread
```
Smaller range / IQR(Q75 − Q25) / P90–P10(spread of most teams) -> tighter distribution -> more parity

```{r}
cutoff_df <- df_top |>
  group_by(league) |>
  summarise(
    n_teams = n(),
    playoff_line = n_teams - 16 + 0.5)
cutoff_df

ggplot(df_top, aes(x = WinPerc, y = reorder(team, WinPerc), color = top3)) +
  geom_point(size = 3) +
  geom_hline(
    data = cutoff_df,
    aes(yintercept = playoff_line),
    linetype = "dashed",
    color = "blue",
    inherit.aes = FALSE) +
  facet_wrap(~league, scales = "free_y") +
  scale_color_manual(values = c("black", "red")) +
  labs(
    title = "Team Win% in NHL vs KHL (2024–25)",
    x = "Win Percentage",
    y = "Team",
    color = "Top 3") +
  theme_minimal()
```

*Fligner–Killeen Hypothesis Test*

H0: NHL and KHL have the same spread of Win% (same parity)
H1: NHL spread is smaller than KHL (NHL has more parity)
```{r}
fligner.test(WinPerc ~ league, data = df)
```
p ≥ 0.05 -> not enough evidence that spreads differ


*Permutation test on IQR*

```{r}
obs <- with(df, IQR(WinPerc[league=="KHL"]) - IQR(WinPerc[league=="NHL"]))
obs
```

```{r}
set.seed(1)

B <- 10000
diffs <- numeric(B)

for(i in 1:B){
  shuffled <- df |>
    mutate(league_shuf = sample(league))
  
  diffs[i] <- with(shuffled,
                   IQR(WinPerc[league_shuf=="KHL"]) - IQR(WinPerc[league_shuf=="NHL"]))
}

# one-sided p-value for "NHL has smaller spread" = "KHL - NHL is big/positive"
p_val <- mean(diffs >= obs)
p_val
```

```{r}
df_test<- df |>
  group_by(league) |>
  summarise(
    IQR = IQR(WinPerc),
    MAD = median(abs(WinPerc - median(WinPerc))),
    P90_P10 = quantile(WinPerc, 0.90) - quantile(WinPerc, 0.10),
    .groups="drop"
  )
df_test
```

In all three measures, the KHL spread is larger than the NHL spread

```{r}
ggplot(df, aes(x = GF_pg, y = WinPerc)) +
  geom_point() +
  facet_wrap(~league) +
  labs(
    title = "Offense vs Winning",
    x = "Goals Scored per Game",
    y = "Win Percentage"
  ) +
  theme_minimal()
```

```{r}
ggplot(df, aes(x = GA_pg, y = WinPerc)) +
  geom_point() +
  facet_wrap(~league) +
  labs(
    title = "Defense vs Winning",
    x = "Goals Allowed per Game",
    y = "Win Percentage"
  ) +
  theme_minimal()
```

```{r}
df |>
  group_by(league) |>
  summarise(corr_offense = cor(GF_pg, WinPerc),
            corr_defense = cor(GA_pg, WinPerc))
```

Offense matters more in KHL, while defense matters more in NHL
