---
title: "Doughnut Run: Time Adjustments & Performance"
format: html
---

In some races (especially fun runs), runners may stop for bonus challenges—like eating doughnuts. This worksheet explores:

**How many doughnuts did runners eat, and how is that related to race time?**

You will work with three datasets:

- `doughnut2015.csv`: adjusted race results
- `doughnut2015unadj.csv`: unadjusted race results
- `doughnuttime.csv`: the “bonus time thresholds” for each doughnut

You’ll practice realistic data analyst tasks:
1. cleaning messy time strings
2. converting time into seconds for analysis
3. joining datasets by an ID
4. turning thresholds into a derived variable (donuts eaten)
5. visualizing patterns and fitting a simple model

# 0. Load the Following Packages

```{r}
#| message: false
library(tidyverse)
library(lubridate)
```


# 1. Load in the CSV Files

Load the three datasets from your data folder.

```{r}
doughnut2015     <- read_csv("Tim_Kashkarov/doughnut2015.csv")
doughnut2015unadj<- read_csv("Tim_Kashkarov/doughnut2015unadj.csv")
doughnuttime     <- read_csv("Tim_Kashkarov/doughnuttime.csv")
```


# 2. Fix the Time Column Format

In these datasets, the Time column may be inconsistent. Some values look like mm:ss:00 but should really be interpreted as 00:mm:ss.

We will make sure every time is in HH:MM:SS format.

Below is a function fix_time() that:
1. turns mm:ss:00 into 00:mm:ss
2. pads h:m:s into hh:mm:ss
3. returns NA if the format is not usable

You don't need to write it again.

```{r}
fix_time <- function(x) {
  x <- as.character(x)
  parts <- strsplit(x, ":", fixed = TRUE)

  vapply(parts, function(p) {
    if (length(p) != 3) return(NA_character_)

    # if third element is "00", treat as mm:ss:00 -> 00:mm:ss
    if (p[3] == "00") {
      sprintf("00:%02d:%02d", as.integer(p[1]), as.integer(p[2]))
    } else {
      sprintf("%02d:%02d:%02d", as.integer(p[1]), as.integer(p[2]), as.integer(p[3]))
    }
  }, character(1))
}
```

b) Apply the function to both race result datasets

```{r}

```


# 3. Convert HH:MM:SS to Total Seconds

For analysis, it is easier to compare times as numbers (seconds).

a) Let's have a new column time_sec in each race dataset

```{r}
doughnut2015 <- doughnut2015 |>
  mutate(time_sec = as.numeric(as.difftime(Time, format = "%H:%M:%S", units = "secs")))

doughnut2015unadj <- doughnut2015unadj |>
  mutate(time_sec = as.numeric(as.difftime(Time, format = "%H:%M:%S", units = "secs")))
```


# 4. Join Adjusted and Unadjusted Results + Compute Time Difference

You want one dataset where each runner has:

- adjusted time (sec)
- unadjusted time (sec)
- difference: unadjusted - adjusted

a) Create a dataset df_diff using an inner join by Race Number
Tip: Rename before joining so you avoid suffix + rename steps.

```{r}

```


# 5. Convert Bonus Thresholds to Seconds + Compute Doughnuts Eaten

The dataset doughnuttime contains a time threshold for each doughnut in the Bonus column. If a runner’s time difference is greater than a threshold, they earned that doughnut.

a) Convert Bonus to seconds

```{r}

```

b) Create a Donut column in df_diff
Use findInterval(): it counts how many thresholds a value exceeds 

```{r}

```


# 6. Explore the Results

a) Make a plot of distribution of doughnuts eaten

```{r}

```

b) Make a plot of doughnuts by finish group

First, create a finish group based on Position using mutate:
1–10, 11–50, 51–100, 101–260

```{r}

```

c) Make a plot showing relationship between doughnuts and adjusted time
```{r}

```

d) Remove non-eaters and re-check the relationship
```{r}

```


# 7. Modeling: Is There an “Optimal” Number of Doughnuts?

Fit a quadratic model:

$time = \beta_0 + \beta_1 Donut + \beta_2 Donut^2$.

Then estimate the “optimal” number of doughnuts (the turning point).
```{r}

```

# 8. Compare Patterns by Gender

Use a scatterplot + smoother, faceted by Gender.
```{r}

```

# 9. Reflection Questions (Write Short Answers)

**1. Do runners who eat more doughnuts tend to be faster or slower? Is it linear?**



**2. Why might the relationship differ after removing runners with 0 doughnuts?**



**3. Based on your quadratic model, does the “optimal” doughnut count make sense?**



**4. What is one limitation of using time_diff_sec as a proxy for doughnuts eaten?**

