---
title: "Doughnut Run: Time Adjustments & Performance"
format: html
---

In some races (especially fun runs), runners may stop for bonus challenges—like eating doughnuts. This worksheet explores:

**How many doughnuts did runners eat, and how is that related to race time?**

You will work with three datasets:

- `doughnut2015.csv`: adjusted race results
- `doughnut2015unadj.csv`: unadjusted race results
- `doughnuttime.csv`: the “bonus time thresholds” for each doughnut

You’ll practice realistic data analyst tasks:
1. cleaning messy time strings
2. converting time into seconds for analysis
3. joining datasets by an ID
4. turning thresholds into a derived variable (donuts eaten)
5. visualizing patterns and fitting a simple model

# 0. Load the Following Packages

```{r}
#| message: false
library(tidyverse)
library(lubridate)
```


# 1. Load in the CSV Files

Load the three datasets from your data folder.

```{r}
doughnut2015     <- read_csv("Tim_Kashkarov/doughnut2015.csv")
doughnut2015unadj<- read_csv("Tim_Kashkarov/doughnut2015unadj.csv")
doughnuttime     <- read_csv("Tim_Kashkarov/doughnuttime.csv")
```


# 2. Fix the Time Column Format

In these datasets, the Time column may be inconsistent. Some values look like mm:ss:00 but should really be interpreted as 00:mm:ss.

Your goal: make sure every time is in HH:MM:SS format.

a) Write a function fix_time() that:
1. turns mm:ss:00 into 00:mm:ss
2. pads h:m:s into hh:mm:ss
3. returns NA if the format is not usable

```{r}
fix_time <- function(x) {
  x <- as.character(x)
  parts <- strsplit(x, ":", fixed = TRUE)

  vapply(parts, function(p) {
    if (length(p) != 3) return(NA_character_)

    # if third element is "00", treat as mm:ss:00 -> 00:mm:ss
    if (p[3] == "00") {
      sprintf("00:%02d:%02d", as.integer(p[1]), as.integer(p[2]))
    } else {
      sprintf("%02d:%02d:%02d", as.integer(p[1]), as.integer(p[2]), as.integer(p[3]))
    }
  }, character(1))
}
```

b) Apply the function to both race result datasets

```{r}
doughnut2015 <- doughnut2015 |>
  mutate(Time = fix_time(Time))

doughnut2015unadj <- doughnut2015unadj |>
  mutate(Time = fix_time(Time))
```


# 3. Convert HH:MM:SS to Total Seconds

For analysis, it is easier to compare times as numbers (seconds).

a) Create a new column time_sec in each race dataset

```{r}
doughnut2015 <- doughnut2015 |>
  mutate(time_sec = as.numeric(as.difftime(Time, format = "%H:%M:%S", units = "secs")))

doughnut2015unadj <- doughnut2015unadj |>
  mutate(time_sec = as.numeric(as.difftime(Time, format = "%H:%M:%S", units = "secs")))
```


# 4. Join Adjusted and Unadjusted Results + Compute Time Difference

You want one dataset where each runner has:

- adjusted time (sec)
- unadjusted time (sec)
- difference: unadjusted - adjusted

a) Create a dataset df_diff using an inner join by Race Number
Tip: Rename before joining so you avoid suffix + rename steps.

```{r}
df_diff <- doughnut2015 |>
  transmute(
    `Race Number`,
    time_sec_adjusted = time_sec,
    Position,
    Gender) |>
  inner_join(
    doughnut2015unadj |>
      transmute(`Race Number`, time_sec_unadjusted = time_sec),
    by = "Race Number") |>
  mutate(time_diff_sec = time_sec_unadjusted - time_sec_adjusted)
```


# 5. Convert Bonus Thresholds to Seconds + Compute Doughnuts Eaten

The dataset doughnuttime contains a time threshold for each doughnut in the Bonus column. If a runner’s time difference is greater than a threshold, they earned that doughnut.

a) Convert Bonus to seconds

```{r}
doughnuttime <- doughnuttime |>
  mutate(Bonus_sec = as.numeric(hms(Bonus)))
```

b) Create a Donut column in df_diff
Use findInterval(): it counts how many thresholds a value exceeds 

```{r}
df_diff <- df_diff |>
  mutate(Donut = findInterval(time_diff_sec, sort(doughnuttime$Bonus_sec)))
```


# 6. Explore the Results

a) Make a plot of distribution of doughnuts eaten

```{r}
df_diff |>
  ggplot(aes(x = Donut)) +
  geom_bar(fill = "#d95f02") +
  labs(title = "Distribution of Doughnuts Eaten",
       x = "Number of Doughnuts",
       y = "Number of Runners") +
  theme_minimal() 
```

b) Make a plot of doughnuts by finish group

First, create a finish group based on Position using mutate:
1–10, 11–50, 51–100, 101–260

```{r}
df_diff <- df_diff |>
  mutate(rank_group = cut(Position, breaks = c(0, 10, 50, 100, 260)))

df_diff |>
  ggplot(aes(x = rank_group, y = Donut)) +
  geom_boxplot(fill = "#d95f02") +
  labs(
    title = "Doughnuts by Finish Group",
    x = "Finish Group",
    y = "Doughnuts"
  ) +
  theme_minimal() 
```

c) Make a plot showing relationship between doughnuts and adjusted time
```{r}
df_diff |>
  ggplot(aes(x = Donut, y = time_sec_adjusted)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Adjusted Time vs Doughnuts Eaten",
    x = "Doughnuts",
    y = "Adjusted Time (sec)"
  ) +
  theme_minimal() 
```

d) Remove non-eaters and re-check the relationship
```{r}
df_diff2 <- df_diff |>
  filter(Donut > 0)

df_diff2 |>
  ggplot(aes(x = Donut, y = time_sec_adjusted)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Adjusted Time vs Doughnuts (Only Runners Who Ate ≥ 1)",
    x = "Doughnuts",
    y = "Adjusted Time (sec)"
  ) +
  theme_minimal() 
```


# 7. Modeling: Is There an “Optimal” Number of Doughnuts?

Fit a quadratic model:

$time = \beta_0 + \beta_1 Donut + \beta_2 Donut^2$.

Then estimate the “optimal” number of doughnuts (the turning point).
```{r}
quad_model <- lm(time_sec_adjusted ~ Donut + I(Donut^2), data = df_diff2)
summary(quad_model)

b <- coef(quad_model)
opt_donuts <- -b["Donut"] / (2 * b["I(Donut^2)"])
opt_donuts
```

# 8. Compare Patterns by Gender

Use a scatterplot + smoother, faceted by Gender.
```{r}
df_diff |>
  ggplot(aes(x = Donut, y = time_sec_adjusted)) +
  geom_point() +
  geom_smooth() +
  geom_smooth(method = lm, formula = y ~ poly(x, 2)) +
  facet_wrap(~Gender, ncol = 1) +
  labs(
    title = "Adjusted Time vs Doughnuts, by Gender",
    x = "Doughnuts",
    y = "Adjusted Time (sec)"
  ) +
  theme_minimal() 
```

# 9. Reflection Questions (Write Short Answers)

1. Do runners who eat more doughnuts tend to be faster or slower? Is it linear?

Runners who eat more doughnuts do not consistently get faster or slower in a straight-line way. The relationship is not linear. The scatterplot suggests a curved pattern: runners who eat a moderate number of doughnuts tend to have faster adjusted times, but very high doughnut counts tend to be associated with slower times.

2. Why might the relationship differ after removing runners with 0 doughnuts?

The group with 0 doughnuts likely includes very different types of runners. Some may be fast runners who skipped the doughnut stops, while others may be slower runners who did not participate. Because the 0-doughnut group mixes different behaviors, it can distort the pattern. Removing them focuses only on runners who actually ate doughnuts, making the relationship clearer.

3. Based on your quadratic model, does the “optimal” doughnut count make sense?

The quadratic model predicts an “optimal” value of about 12.7 doughnuts, meaning the fastest adjusted time occurs around 13 doughnuts. While this makes sense mathematically for a U-shaped curve, it may not be realistic because very few runners ate that many doughnuts.

4. What is one limitation of using time_diff_sec as a proxy for doughnuts eaten?

A limitation is that time_diff_sec may include more than just time spent eating doughnuts. It could also include other stopping time (resting, water breaks, congestion, timing noise, etc.). Therefore, the time difference is only an approximation and may not perfectly represent doughnuts eaten.
