---
title: "Understanding Nathan's Hot Dog Contest"
author: Amara McGowan
format:
  closeread-html:
    cr-style:
      narrative-background-color-sidebar: darkgreen
      section-background-color: white
      narrative-text-color-sidebar: white
---

```{r}
#| include: false
library(tidyverse)
hotdog_winners <- read_csv("HotdogData.csv")
```

```{r}
#| include: false
hotdogs_per_min <- hotdog_winners |> mutate(hotdogs_per_minute = Dogs/Time)
```

:::{.cr-section}

There's a lot of money behind the first place title of Nathan's Hot Dog Eating Contest. So, how many hotdogs does someone really have to eat to win the contest, and has it always been that way? Do competitive eaters just get better and better every year, eating more and more hotdogs? 

Let's find out. 

Here is a plot of how many hotdogs the winner of Nathan's Hot Dog Eating Contest ate each year. @cr-plot1

:::{#cr-plot1}
```{r}
ggplot(data = hotdog_winners, aes(x = Year, y = Dogs)) +
  geom_point() +
  theme_minimal() +
  labs(y = "Hotdogs Consumed", title = "Amount of Hotdogs Consumed by Winners")
```
:::

Hmmm, this plot looks pretty interesting. Why is there a dip in the middle and how come people can eat so many more hotdogs now than they used to be able to?

Thinking about our variables, the duration of the contest (Time) is not similar across different years, so it is not really fair to compare winners from earlier years to more recent years based on total dogs eaten.

When we make a new variable called hotdogs_per_minute, we are able to see how many hotdogs each winner was able to eat per minute instead of overall, helping us see who the best eaters really were without being distracted by who had the longest time to eat. @cr-plot2

:::{#cr-plot2}
```{r}
ggplot(data = hotdogs_per_min, aes(x = Year, y = hotdogs_per_minute)) +
  geom_point() +
  theme_minimal() +
  labs(y = "Hotdogs Consumed per min", title = "Amount of Hotdogs Consumed per by Winners")
```
:::

What's wrong with this plot? The general pattern is still similar. Why might that be? 

One explanation is that the duration of the competition changes. When we look at the competetion in its earlier 5 years compared to its latest five years. @cr-table1

:::{#cr-table1}

```{r}
earliest_five <- hotdogs_per_min |>
  arrange(Year) |>
  slice_head(n = 5)

latest_five <- hotdogs_per_min |>
  arrange(desc(Year)) |>
  slice_head(n = 5)

table <- bind_rows(earliest_five, latest_five) |>
  arrange(Year)

table
```
:::

The earliest five winners competed in competitions all under 5 minutes long, while the most recent five winners, some years repeating, are 10 minutes long. This means in earlier years they were eating for less time and therefore may have been able to eat more for longer, whereas the more recent competitors may have slowed down after the first few minutes of eating, lowering their average hotdog per minute rate. 

That still doesn't explain why there is a clear jump around the early 2000s. @cr-plot2

What happened there? Let's find out who won that year! @cr-table2

:::{#cr-table2}
```{r}
hotdogs_per_min |>
  filter(Year > 1998) |>
  arrange((Year)) |>
  slice_head(n = 5)
```

:::




Again, another new one. @cr-plot3

:::{#cr-plot3}
```{r}
ggplot(data = hotdogs_per_min, aes(x = Year, y = hotdogs_per_minute, color = Sex)) +
  geom_point() +
  theme_minimal() +
  labs(y = "Hotdogs Consumed per min", title = "Amount of Hotdogs Consumed per by Winners")
```
:::

:::