---
title: "Looking into Nathan's Famous Hot Dog Eating Contest with Closeread"
author: Amara McGowan
format:
  closeread-html:
    cr-style:
      narrative-background-color-sidebar: darkgreen
      section-background-color: transparent
      narrative-text-color-overlay: black
      narrative-text-color-sidebar: white
---

```{r}
#| include: false
library(readr)
library(tidyverse)
```

```{r}
#| include: false
#hotdog_clean <- read_csv("Amara-McGowan/hotdog-data-storytelling-module/hotdog_clean.csv")
hotdog_clean <- read_csv("hotdog_clean.csv")

hotdog_men <- hotdog_clean |> filter(Sex == 1)
hotdog_women <- hotdog_clean |> filter(Sex == 0)

hotdog_men_min <- hotdog_men |> mutate(Hotdogs_per_minute = Dogs/Time)
hotdog_women_min <-hotdog_women |> mutate(Hotdogs_per_minute = Dogs/Time)

ggplot(data = hotdog_men_min, aes(x = Year, y = Hotdogs_per_minute)) +
  geom_point() +
  theme_minimal() +
  labs(y = "Hotdogs Per Minute", title = "Amount of Hotdogs Consumed Per Minute by Winners")
```

:::{#cr-intro}

Nathan's Famous Hot Dog Eating Contest is a competitive eating competition where contestants are given 10 minutes to eat as many hotdogs (bun and all) as they can manage. In its early years the duration of the competition varied, but for the past decade it has been a 10-minute cutoff.

If you're interested in understanding how competitive eaters are able to consume so much food, here is a video referencing Nathan's Famous Hot Dog Eating Contest and the science behind it. 

<iframe width="560" height="315" src="https://www.youtube.com/embed/D9DZmoRY0H4?si=AVxTT6zn5CAQxNfR" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe> 

::: {.callout-note collapse="true" title="Learning Objectives" appearance="minimal"}
By the end of this tutorial, students will be able to:

1. Create a "scrollytelling" document with Closeread.

2. Create scatterplots to look at patterns in data.

3. Form questions about the data based on visualizations. 

4. Investigate questions about the data.
:::

::: {.callout-note collapse="true" title="Activity Length" appearance="minimal"}
This could be suitable of an out of class activity or one that spans one
or more class periods.
:::

::: {.callout-note collapse="true" title="Methods" appearance="minimal"}
**Technology Requirement:**

This activity is designed to be completed primarily in `R` using a
Quarto document workflow. Students should be comfortable importing data
and making basic plots with packages from the tidyverse
(especially `readr` and `ggplot2`).
:::

::: {.callout-caution collapse="true" title="Research Questions" appearance="minimal"}

During this lab, you'll investigate the following research questions:

+ Question one
+ Question two

:::


### Data

The `hotdog_clean.csv` dataset contains 69 rows, one per
Nathan's Famous Hot Dog Eating Contest Winner, with returning winners
showing up in the data multiple times. The primary variables in the 
data are `Dogs` and `Time`.

Download Data:
<!--TODO: [add later](add later) -->

<details>

<summary><b>Variable Descriptions</b></summary>

| Variable | Description |
|------------------------------------|------------------------------------|
| Year | Year the Competition took place |
| Sex | Sex of the Winner (0 = Woman, 1 = Man) |
| Dogs | How many hotdogs the winner consumed |
| Winner | First and Last Name of the Winner |
| Time | Duration of the competition in minutes |

</details>

#### Data Sources

-   [Wikipedia.org](Wikipedia.org/wiki/Nathan%27s_Hot_Dog_Eating_Contest#By_year_(color-coded_by_belt_color)){target="_blank"}

::: @cr-intro

::::: {.cr-section}

:::{#cr-tibble1 .cr-sticky}
```{r}
#| warning: false
head(hotdog_clean)
```
:::

This module will help you learn how to use the quarto extension `closeread` to tell stories with your data. To the right is a plot of the amount of hotdogs consumed per minute by winners. When you scroll, you'll see the plot stays to the right and new text emerges. @cr-tibble1

This scrolling feature helps you tell stories about your data, some people call it "scrollytelling." @cr-tibble1
:::

::::: {#cr-main}

```{r, warning = FALSE}
library(readr)
library(tidyverse)
```

Here. @cr-main 

::::: @cr-main
