---
title: "An Introduction to Data Scrollytelling: Nathan's Hot Dog Eating Contest"
author: 
- name: Amara McGowan
  url: https://www.linkedin.com/in/amaramcgowan/
format: html
categories:
  - Data Visualization
  - Closeread Extention
  - Data Storytelling
---

Please note that these material have not yet completed the required
pedagogical and industry peer-reviews to become a published module on
the SCORE Network. However, instructors are still welcome to use these
materials if they are so inclined.

# Welcome Video

If you are unfamiliar with Nathan's Hot Dog Contest, please watch the first minute of this video:

<iframe width = "560" height = "315" src = "https://www.youtube.com/embed/txMPjvIlVpk?si=hXjLB4LBoAESAx1t&end=75" title = "YouTube video player" frameborder = "0" allow = "accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy = "strict-origin-when-cross-origin" allowfullscreen></iframe>

# Introduction 

Nathan's Famous, an American hot dog brand established in Brooklyn, New York, in 1916, is widely recognized for its annual hot dog eating contest held each year on July 4th. Originally created as a promotional event to increase brand visibility, the contest has evolved into what many call the "Super Bowl" of competitive eating. Beginning in 2008, the event was standardized at a duration of ten minutes, though earlier competitions ranged from 2.5 to 12 minutes long. In order to compete, participants must be at least 18 years old and qualify by winning a smaller regional contest. On the day of the event, between 15 and 30 competitive eaters participate. In 2011, the organizers introduced separate divisions for men and women, resulting in two first-place winners each year. The top five competitors in each division receive cash prizes, with first place earning $10,000 in 2025.  

In this module, you will learn how to use the quarto extension `closeread` to tell a story using data from Nathan's Hot Dog Contest. Upon completion of this module, you will have a Closeread document that looks similar to this: [Example Closeread](TODO: Fix Link!){target="_blank"}

::: {.callout-note collapse = "true" title = "Learning Objectives" appearance = "minimal"}
By the end of this tutorial, you will be able to:

1.  Identify meaningful questions about a dataset.

2.  Transform a dataset to answer those questions. 

3.  Use the `Closeread` Quarto extension to create a visually engaging story.
:::

::: {.callout-note collapse = "true" title = "Activity Length" appearance = "minimal"}
This could be suitable of an out of class activity or one that spans one
or more class periods.
:::

::: {.callout-note collapse = "true" title = "Methods" appearance = "minimal"}
**Technology Requirement:**

This activity is designed to be completed primarily in `R` using a
Quarto document workflow with a Closeread format. Students should be comfortable importing data and
making basic plots with packages from the tidyverse
(especially `readr`, `dplr`, and `ggplot2`).

No previous experience with Closeread is required.
:::

# Data

The `HotdogData.csv` dataset contains 69 rows, each representing a winner of Nathan's Hot Dog Eating Contest. Returning winners appear multiple times, and many years include two winners (one man and one woman). The primary variables in the dataset are `Dogs` and `Time`.

Download Data: [HotdogData.csv](HotdogData.csv)

<details>
<summary><b>Variable Descriptions</b></summary>

| Variable | Description |
|------------------------------------|------------------------------------|
| Year | Year the competition took place |
| Sex | Sex of the winner (0 = Woman, 1 = Man) |
| Dogs | Total hot dogs consumed by the winner |
| Winner | First and Last Name of the Winner |
| Time | Duration of the competition in minutes |
</details>

<details>
<summary><b>Data Source</b></summary>
The data was scraped and cleaned from this Wikipedia page:

[https://en.wikipedia.org/wiki/Nathan's_Hot_Dog_Eating_Contest#By_year_(color-coded_by_belt_color)](https://en.wikipedia.org/wiki/Nathan's_Hot_Dog_Eating_Contest#By_year_(color-coded_by_belt_color)){target="_blank"}
</details>

# Installing Closeread

Follow along below or refer to the [Closeread homepage](https://closeread.dev) for more information about installation. 

1. In RStudio, open the project folder where you'd like to work.

2. At the bottom of the RStudio window, locate the Console. Next to it, select the Terminal tab. 

3. In the terminal, paste and run the following command:

```{r echo = TRUE, eval = FALSE}
quarto add qmd-lab/closeread
```

4. When prompted with messages like <i>"Do you trust the authors of this extension?"</i>  or <i>"Would you like to continue?"</i>, type Y into the terminal to confirm and proceed with the installation. 

5. After completing these steps, the Quarto extension Closeread will be installed. 

# Using Closeread

Before we get started on a Quarto document, please take a moment to read through this section to learn more about how to use Closeread.

To use the Closeread extension in a Quarto document, you have to include the following in your document header:

```yaml
---
format:
  closeread-html:
---
```

You can customize your Closeread with `cr-style`, which you include in your header. 

<details>
<summary><b>Click to reveal some basic customization options</b></summary>

```yaml
---
format:
  closeread-html: 
    cr-style:
  
      # Sidebar Background (text area as you scroll)
      narrative-background-color-sidebar: darkgreen
      
      # Section Background (area next to the sidebar)
      section-background-color: white
      
      # Sidebar Text Color
      narrative-text-color-sidebar: white
---
```
</details>


If you want to customize your Closeread further, there are more options here: [Closeread Styling Page](https://closeread.dev/guide/styling.html)

To make a Closeread document, you need a section, a sticky, and a trigger. 

::: {.callout-note collapse = "true" title = "Section" appearance = "minimal"}
A section (`cr-section`) enables Closeread features. If you type inside of the section, the text will appear on your sidebar. If you type outside of your section, the text will render like it would on a normal Quarto document. This is how you format a section:

```markdown

Anything you put here would render normally 

:::{.cr-section} 

You can put graphs, text, and pictures here. 

When you are typing the text for your sidebars, this is where you put it.

This is also where you will put your stickies!

::: 
```
:::

::: {.callout-note collapse = "true" title = "Sticky" appearance = "minimal"}
Stickies (e.g. `#cr-ID`) stay stuck to the page, fixed in place while your sidebar text changes. You can name the ID part of the label anything you'd like, for example #cr-HotdogPlot or #cr-table. This is how you format a sticky:

```markdown

:::{.cr-section} 

:::{#cr-ID}

# You can put a code chunk with a plot here

:::

:::
```
:::

::: {.callout-note collapse = "true" title = "Trigger" appearance = "minimal"}
A trigger (e.g. `@cr-ID`) allows you to make something from the section, like text, be the signal that a particular graph should appear on the right of the sidebar at the same time as whatever text the trigger is on. This is how you format a trigger:

```markdown

:::{.cr-section}

This text would appear at the same time as my plot. @cr-myplot

:::{#cr-myplot}

# You would put the code chunk for my plot in here

:::

:::

```
:::

If you would like more information about the three parts of a Closeread (the section, stickies, and triggers) you can review this page: [Closeread Components Guide](https://closeread.dev/guide/components.html)


# Getting Started 

Create a new Quarto document in the project folder where you installed the Closeread extension. 

::: {.callout-note  title = "Exercise 1: Enable Closeread & Load Data"}
a.  Try editing the YAML header to enable Closeread and customize the colors. 

<details>
<summary><b>Click for solution</b></summary>

```yaml
---
title: "Your title"
author: Your name
format:
  closeread-html:
    cr-style:
      narrative-background-color-sidebar: darkgreen
      section-background-color: transparent
      narrative-text-color-sidebar: white
---
```
</details>

b.  Download [HotdogData.csv](HotdogData.csv) to your project folder. Then, Load tidyverse and read `HotdogData.csv` as `hotdog_winners`, making sure that the code will not appear when you render the document.

<details>
<summary><b>Click for solution</b></summary>

```{r}
#| include: false

library(tidyverse)
hotdog_winners <- read_csv("HotdogData.csv")
```

```markdown
#| include: false

library(tidyverse)

hotdog_winners <- read_csv("HotdogData.csv")
```
</details>

:::

::: {.callout-note  title = "Exercise 2: Brainstorm Questions"}
Now our document is ready! What story should we tell? 

Think of 2 questions that you would like to investigate about these winners.

<details>
<summary><b>Click for example questions</b></summary>
Some questions we might explore:

1.  How many hotdogs does someone need to eat to win the contest and has that number gone up over time?

2.  Do male and female winners show different patterns over time?  
</details>
:::


::: {.callout-note  title = "Exercise 3: Start Your Closeread Story"}
For our story, we will investigate: "How many hotdogs does someone need to eat to win the contest and has that number gone up over time?" 

Create a `cr-section`. Inside, introduce your reader to Nathan's contest and your guiding questions with plain text. 

<details>
<summary><b>Click for an example</b></summary>

```markdown

Code to read in the data

:::{.cr-section}

Nathan's Hot Dog Eating Contest is a competitive eating contest where people from around
the world try to eat as many hotdogs as possible in a limited amount of time. The contest
has not always had a set duration, but the competition has been fixed at 10 minutes long since
2008. The top five competitors of the contest are awarded large cash prizes, with the 1st place 
prize in 2025 being $10,000. 10 grand is a lot of money. How much does it really take to win 1st
place in Nathan's Hot Dog Eating Contest? Does it get harder and harder every year? In other 
words, how many hotdogs do the winners eat and do the winners eat more and more hotdogs every year?

:::
```
</details>
:::

Now that we have introduced the readers of our story to our data and guiding questions, we should make a plot to start working on answering our questions.

When you are data scrollytelling with Closeread, it is nice to put visuals on the right side of the sidebar in stickies, so that readers can look at the visuals while they scroll through text that explains what they are looking at.

Let's make a sticky that will help us start to share with our readers how many hotdogs winners have to eat to claim first place and whether the number of hotdogs they have to eat to win has increased over the years. 

::: {.callout-note  title = "Exercise 4: Create Your First Sticky"}
a.  Create a sticky called `#cr-plot1` inside your `cr-section`.

<details>
<summary><b>Click for solution</b></summary>

```markdown

Code to read in the data

:::{.cr-section}

Your introduction text here...

:::{#cr-plot1}

:::

:::
```
</details>

b.  Inside the sticky, plot Total Hotdogs Eaten by Winners Over Time. 

<details>
<summary><b>Click for solution</b></summary>

```{r}
ggplot(data = hotdog_winners, aes(x = Year, y = Dogs)) +
  geom_point() +
  theme_minimal() +
  labs(y = "Hotdogs Eaten", title = "Total Hotdogs Eaten by Winners Over Time")
```

</details>

c.  Add a trigger (`@cr-plot1`) to make your sticky appear.   

    Write 2-3 sentences reacting to the plot. What patterns do you notice? What seems surprising? If        your sentences are spaced out, add your trigger to the first sentence and your sticky will be stuck     to the right side of the screen until you trigger another sticky or end the section. 

<details>
<summary><b>Click for example </b></summary>

```markdown

Code to read in the data

:::{.cr-section}

Your intro text here...

:::{#cr-plot1}

Code for plot...

:::

The plot to the right shows the total hotdogs eaten by every winner over time.
Look at the pattern in the plot. As time goes on, winners do tend to eat more
hotdogs, but there is a weird spike around 2000. Let's find out what happened
to cause the amount of hotdogs winners ate to increase by so much! @cr-plot1

:::
```
</details>
:::

Looking at the plot, did you notice the dramatic jump from 2000 to 2001? It looks like the amount of hotdogs someone has to eat to claim first place nearly doubles in years after 2000. 

Let's figure out why that happened.  

::: {.callout-note  title = "Exercise 5: Create Another Sticky"}

a.  Create a second sticky called `#cr-table1`.

<details>
<summary><b>Click for solution </b></summary>

```markdown

Code to read in the data

:::{.cr-section}

Your intro text here...

:::{#cr-plot1}

Code for plot...

:::

Trigger text here... @cr-plot1

:::{#cr-table1}

Code for table...

:::

:::
```
</details>

b.  Inside your second sticky, create a table showing the data from the years 1999 to 2003.

<details>
<summary><b>Click for solution</b></summary>

```{r}
hotdog_winners |>
  filter(Year >= 1999 & Year <= 2003) |>
  arrange(Year)
```

</details>

c.  Create a trigger (`@cr-table1`) for your second sticky. 

    Write 2-3 sentences reacting to the table. How many hotdogs did the winner in 2000 eat? How many        hotdogs did the winner from 2001 eat? Was it a big difference? Who is the winner from 2001?             Remember, you're telling the reader a story.

<details>
<summary><b>Click for example</b></summary>

```markdown

Taking a closer look at the competitions from 1999 to 2003 shows us who is responsible for the spike in hotdogs consumed in the early 2000s. @cr-table1

In 2000, eating 25.1 hotdogs was enough to claim first place. In 2001, Takeru Kobayashi ate 50, about double the amount of hotdogs eaten by the winner from the year before. How was Kobayashi able to eat double the amount of hotdogs of the previous winner? 

```
</details>
:::

Now that we know Takeru Kobayashi is the winner behind the huge spike in the amount of hotdogs eaten during the competition, let's look him up and find out why he was able to eat so many hotdogs! 

Take a moment to look Kobayashi up and try to figure out how he was able to eat so many more hotdogs than the winner from the year before him. 

If you need help finding out how, try searching the internet for things like <i>"How does Kobayashi eat so many hotdogs?"</i> or <i>"What did Kobayashi do in 2001 to win Nathan's Hot Dog Contest?"</i>


# Closing Video

If you're interested in understanding how competitive eaters are able to consume so much food, here is a video referencing Nathan's Famous Hot Dog Eating Contest and the science behind it:

<iframe width="560" height="315" src="https://www.youtube.com/embed/D9DZmoRY0H4?si=AVxTT6zn5CAQxNfR" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe> 

# References

Bray, A., & Goldie, J. (2024). *Closeread*. https://closeread.dev/