---
title: "An Introduction to Data Scrollytelling: Nathan's Hot Dog Eating Contest"
author: 
- name: Amara McGowan
- url: https://www.linkedin.com/in/amaramcgowan/
format: html
categories:
  - Data Visualization
  - Closeread Extention
  - Data Storytelling
---

Please note that these material have not yet completed the required
pedagogical and industry peer-reviews to become a published module on
the SCORE Network. However, instructors are still welcome to use these
materials if they are so inclined.

# Welcome Video

If you are unfamiliar with Nathan's Hot Dog Contest, please watch the first minute of this video:

<iframe width = "560" height = "315" src = "https://www.youtube.com/embed/txMPjvIlVpk?si=hXjLB4LBoAESAx1t&end=75" title = "YouTube video player" frameborder = "0" allow = "accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy = "strict-origin-when-cross-origin" allowfullscreen></iframe>

# Introduction 

Nathan's Famous, an American hot dog brand established in Brooklyn, New York, in 1916, is widely recognized for its annual hot dog eating contest held each year on July 4th. Originally created as a promotional event to increase brand visibility, the contest has evolved into what many call the "Super Bowl" of competitive eating. Beginning in 2008, the event was standardized at a duration of ten minutes, though earlier competitions ranged from 2.5 to 12 minutes long. In order to compete, participants must be at least 18 years old and qualify by winning a smaller regional contest. On the day of the event, between 15 and 30 competitive eaters participate. In 2011, the organizers introduced separate divisions for men and women, resulting in two first-place winners each year. The top five competitors in each division receive cash prizes, with first place earning $10,000 in 2025.  

In this module, you will learn how to use the quarto extension `closeread` to tell a story using data from Nathan's Hot Dog Contest. TODO: add example!

::: {.callout-note collapse = "true" title = "Learning Objectives" appearance = "minimal"}
By the end of this tutorial, you will be able to:

1.  Identify meaningful questions about a dataset.

2.  Transform a dataset to answer those questions. 

3.  Use the `Closeread` Quarto extension to create a visually engaging story showcasing your findings.
:::

::: {.callout-note collapse = "true" title = "Activity Length" appearance = "minimal"}
This could be suitable of an out of class activity or one that spans one
or more class periods.
:::

::: {.callout-note collapse = "true" title = "Methods" appearance = "minimal"}
**Technology Requirement:**

This activity is designed to be completed primarily in `R` using a
Quarto document workflow with a Closeread format. Students should be comfortable importing data and
making basic plots with packages from the tidyverse
(especially `readr`, `dplr`, and `ggplot2`).

No previous experience with Closeread is required.
:::

# Data

The `HotdogData.csv` dataset contains 69 rows, each representing a winner of Nathan's Hot Dog Eating Contest. Returning winners appear multiple times, and many years include two winners (one man and one woman). The primary variables in the dataset are `Dogs` and `Time`.

Download Data: [HotdogData.csv](HotdogData.csv)

<details>
<summary><b>Variable Descriptions</b></summary>

| Variable | Description |
|------------------------------------|------------------------------------|
| Year | Year the competition took place |
| Sex | Sex of the winner (0 = Woman, 1 = Man) |
| Dogs | Total hot dogs consumed by the winner |
| Winner | First and Last Name of the Winner |
| Time | Duration of the competition in minutes |
</details>

<details>
<summary><b>Data Source</b></summary>
The data was scraped and cleaned from this Wikipedia page:

[https://en.wikipedia.org/wiki/Nathan's_Hot_Dog_Eating_Contest#By_year_(color-coded_by_belt_color)](https://en.wikipedia.org/wiki/Nathan's_Hot_Dog_Eating_Contest#By_year_(color-coded_by_belt_color)){target="_blank"}
</details>

# Installing Closeread

Follow along below or refer to the [Closeread homepage](https://closeread.dev) for more information about installation. 

1. In RStudio, open the project folder where you'd like to work.

2. At the bottom of the RStudio window, locate the Console. Next to it, select the Terminal tab. 

3. In the terminal, paste and run the following command:

```{r echo = TRUE, eval = FALSE}
quarto add qmd-lab/closeread
```

4. When prompted with messages like <i>"Do you trust the authors of this extension?"</i>  or <i>"Would you like to continue?"</i>, type Y into the terminal to confirm and proceed with the installation. 

5. After completing these steps, the Quarto extension Closeread will be installed. 

# Using Closeread

Before we get started on a Quarto document, please take a moment to read through this section to learn more about how to use Closeread.

To use the Closeread extension in a Quarto document, you have to include the following in your document header:

```yaml
---
format:
  closeread-html:
---
```

You can customize your Closeread with `cr-style`, which you include in your header. 

<details>
<summary><b>Click to reveal some basic customization options</b></summary>

```yaml
---
format:
  closeread-html: 
    cr-style:
  
      # Sidebar Background (text area as you scroll)
      narrative-background-color-sidebar: darkgreen
      
      # Section Background (area next to the sidebar)
      section-background-color: white
      
      # Sidebar Text Color
      narrative-text-color-sidebar: white
---
```
</details>


If you want to customize your Closeread further, there are more options here: [Closeread Styling Page](https://closeread.dev/guide/styling.html)

To make a Closeread document, you need a section, a sticky, and a trigger. 

::: {.callout-note collapse = "true" title = "Section" appearance = "minimal"}
A section (`cr-section`) enables Closeread features. If you type inside of the section, the text will appear on your sidebar. If you type outside of your section, the text will render like it would on a normal Quarto document. This is how you format a section:

```markdown

Anything you put here would render normally 

:::{.cr-section} 

You can put graphs, text, and pictures here. 

When you are typing the text for your sidebars, this is where you put it.

This is also where you will put your stickies!

::: 
```
:::

::: {.callout-note collapse = "true" title = "Sticky" appearance = "minimal"}
Stickies (e.g. `#cr-ID`) stay stuck to the page, fixed in place while your sidebar text changes. You can name the ID part of the label anything you'd like, for example #cr-HotdogPlot or #cr-table. This is how you format a sticky:

```markdown

:::{.cr-section} 

:::{#cr-ID}

# You can put a code chunk with a plot here

:::

:::
```
:::

::: {.callout-note collapse = "true" title = "Trigger" appearance = "minimal"}
A trigger (e.g. `@cr-ID`) allows you to make something from the section, like text, be the signal that a particular graph should appear on the right of the sidebar at the same time as whatever text the trigger is on. This is how you format a trigger:

```markdown

:::{.cr-section}

This text would appear at the same time as my plot. @cr-myplot

:::{#cr-myplot}

# You would put the code chunk for my plot in here

:::

:::

```
:::

If you would like more information about the three parts of a Closeread (the section, stickies, and triggers) you can review this page: [Closeread Components Guide](https://closeread.dev/guide/components.html)


# Getting Started 

To begin, create a new Quarto document in the project folder where you have installed the Closeread extension. Next, edit the document header to enable Closeread. You can customize the colors however you would like, for example:

```yaml
---
title: "Your title"
author: Your name
format:
  closeread-html:
    cr-style:
      narrative-background-color-sidebar: darkgreen
      section-background-color: transparent
      narrative-text-color-sidebar: white
---
```

Next, download the data [HotdogData.csv](HotdogData.csv), get the tidyverse package and read in the dataset using `readr`. 

<details>

<summary><b>Click to reveal code</b></summary>

```{r}
#| echo: true
#| eval: false

library(tidyverse)

hotdog_winners <- read_csv("HotdogData.csv")
```

</details>