---
title: "Hotdog-Story"
format: closeread-html
execute: 
  echo: false
  warning: false
  message: false
---


```{r}
library(tidyverse)
library(readr)

# Load cleaned dataset (your confirmed correct path)
hotdog_clean <- read_csv("../../hotdogs/hotdog_clean.csv")

# Split Genders
hotdog_men <- hotdog_clean %>% filter(Sex == 1)
hotdog_women <- hotdog_clean %>% filter(Sex == 0)

# Standardize: hot dogs per minute
hotdog_men <- hotdog_men %>% mutate(Hotdogs_per_minute = Dogs / Time)
hotdog_women <- hotdog_women %>% mutate(Hotdogs_per_minute = Dogs / Time)
```


:::{.cr-section}
## The Nathan’s Hot Dog Eating Contest

The Nathan’s Famous Hot Dog Eating Contest began in 1972 in Brooklyn, New York.
Over time, rules, competitors, and performance levels have changed dramatically.
:::

:::{.cr-section}

:::{focus-on="cr-rate-hist"}

A fair comparison: hot dogs per minute

To compare winners across time, we standardize performance by using:

Hot dogs per minute = Dogs ÷ Time

Scroll to reveal a histogram of winners’ eating rates (men).
:::

:::{#cr-rate-hist}
```{r}
hist(
  hotdog_men$Hotdogs_per_minute,
  main = "Men’s Winners: Hot Dogs Per Minute",
  xlab = "Hot Dogs per Minute",
  breaks = 12)

```

:::

:::

:::{.cr-section}

:::{focus-on="cr-rate-by-year"}

How has the eating rate changed over time?

Plot hot dogs per minute vs. year

This makes it easier to spot years where performance jumps or trends upward.
:::

:::{#cr-rate-by-year}

```{r}
ggplot(hotdog_men, aes(x = Year, y = Hotdogs_per_minute)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Men’s Winners: Hot Dogs Per Minute by Year",
    x = "Year",
    y = "Hot Dogs per Minute"
  )
```

:::

:::

:::{.cr-section}


