---
title: "Closeread Scrollytelling Module (R + Quarto)"
author:
  - name: Kevin Euyoque
    affiliation:
        name: St. Lawrence University
date: "February 20, 2026"
description: "A data visualization module using Nathan’s Hot Dog Eating Contest data through a closeread story."
editor:
  canonical: true
categories:
  - tidyverse
  - closeread
  - data visualization
  - scrollytelling
---

## Module

Please note that these materials have not yet completed the required pedagogical and industry peer-reviews to become a published module on the SCORE Network. However, instructors are still welcome to use these materials if they are so inclined.


## Introduction

The Nathan’s Famous Hot Dog Eating Contest has been held annually since 1972. Over time, changes in contest duration, training methods, and competitive intensity have shaped modern performance levels.

In this module, students explore contest data through a **closeread scrollytelling format** in Quarto. The goal is not only to analyze performance trends, but to understand how data visualization can be structured into a guided narrative.

Rather than simply producing graphs, students will learn how to structure visualizations into a guided analytical narrative.


## Background: Closeread

[**Closeread**](https://closeread.dev) is a Quarto extension that enables scrollytelling for HTML documents.

Scrollytelling is a style of web-based storytelling in which:

Text and graphics transition as the reader scrolls

Visual elements become “sticky” and remain fixed while the narrative changes

Attention is directed intentionally toward specific analytical moments

Closeread allows these techniques to be implemented directly within Quarto using structured sections and plot references.

### Core Closeread Structure

Closeread documents are built using three key components:

1. Sections

Sections define areas where scrollytelling behavior occurs

```{r}
#| warning: false
'::::{.cr-section}
...
::::'
```

Everything inside a .cr-section behaves as part of the scroll-driven story.


2. Stickies (Plots or Elements to Highlight)

A sticky element is defined with a unique ID:

```{r}
#| warning: false
':::{#cr-plotname}
<plot code>
:::'
```

This element becomes pinned (or “sticky”) when triggered.

3. Referencing a Sticky

Inside the narrative text, reference the sticky using:

@cr-plotname

When that line scrolls into view, the corresponding sticky element becomes the focus.

::: {.callout-note collapse="true" title="Learning Objectives" appearance="minimal"}

By the end of this activity, you will be able to:

1. Create time series visualizations using ggplot
2. Explain why standardizing measurements (ex: hot dogs per minute) improves comparisons across time
3. Identify structural shifts in performance trends
4. Implement closeread syntax to build a scrollytelling data document in Quarto
5. Combine narrative reasoning with visual evidence

:::

::: {.callout-note collapse="true" title="Methods" appearance="minimal"}

Technology Requirements

- R
- RStudio
- Quarto with closeread extension installed

Statistical Concepts Used

- Time series plots
- Standardization (rates vs totals)
- Visual trend analysis
- Annotation of key observations

Technical Concepts Used

- tidyverse for data manipulation and plotting
- Closeread section and sticky syntax
- Narrative-triggered visualization

:::


## Data

This module uses a cleaned dataset of winners from the Nathan’s Famous Hot Dog Eating Contest from 1972 to 2025.

Download data: [hotdog_clean.csv]("hotdog_clean.csv")

<details>

<summary><b>Variable Descriptions</b></summary>

| Variable | Description                                               |
|----------|-----------------------------------------------------------|
| Year     | Year the competition took place                           |
| Sex      | Sex of the winner for the year the competition took place |
| Dogs     | Total hot dogs eaten                                      |
| Winner   | Name of the winner for their sex category                 |
| Time     | Time in minutes the competition elapsed                   |
| DPM      | Dogs eaten per minute                                     |

</details>

## Exercises

Students will:

1. Construct a baseline plot of total hot dogs eaten over time.
2. Identify a structural shift in the early 2000s.
3. Standardize performance using hot dogs per minute.
4. Use closeread references to structure a scroll-based narrative.
5. Highlight Joey Chestnut’s winning years within the larger historical trend.

::: {.callout-note collapse="true" title="Conclusion" appearance="minimal"}

This module demonstrates how statistical reasoning and visual storytelling can be integrated into a cohesive scrollytelling experience. By combining careful metric selection with structured narrative flow, students learn not only how to analyze data, but how to communicate it effectively.

:::

#### Author

Created by Kevin Euyoque 
