---
title: "Closeread Scrollytelling Handout (R + Quarto)"
subtitle: "Nathan’s Famous Hot Dog Eating Contest Module"
author:
  - name: Kevin Euyoque
    affiliation:
      - id: slu
        name: St. Lawrence University
date: "March 2, 2026"
description: "A data visualization module using Nathan’s Hot Dog Eating Contest data through a closeread story."
editor:
  canonical: true
categories:
  - tidyverse
  - closeread
  - data visualization
  - scrollytelling
format:
  closeread-html:
    code-fold: true
    cr-style:
      narrative-background-color-sidebar: "#09753C"
      narrative-text-color-sidebar: white
execute:
  echo: false
  warning: false
  message: false
---

## Module

Please note that these materials have not yet completed the required pedagogical and industry peer-reviews to become a published module on the SCORE Network. However, instructors are still welcome to use these materials if they are so inclined.

---

## What you will need

### Software
- R
- RStudio
- Quarto + the **closeread** extension

### Required packages
- `tidyverse`
- `readr`

::: {.callout-note collapse="true" title="Install packages (if needed)" appearance="minimal"}
```{r}
#| eval: false
install.packages(c("tidyverse", "readr"))
```

:::

## Introduction

The Nathan’s Famous Hot Dog Eating Contest has been held annually since 1972. Over time, changes in contest duration, training methods, and competitive intensity have shaped modern performance levels.

In this module, students explore contest data through a **closeread scrollytelling format** in Quarto. The goal is not only to analyze performance trends, but to understand how data visualization can be structured into a guided narrative.

Rather than simply producing graphs, students will learn how to structure visualizations into a guided analytical narrative.

::: {.callout-note collapse="true" title="Learning Objectives" appearance="minimal"}
By the end of this activity, you will be able to:

- Interpret time-series visualizations of competitive performance.
- Distinguish between totals and standardized rates (e.g., hot dogs per minute).
- Identify structural shifts in long-term trends.
- Use closeread in Quarto to create scrollytelling visualizations.
:::

::: {.callout-note collapse="true" title="Methods" appearance="minimal"}

Technology requirement: This activity is completed in RStudio using Quarto with the closeread format.

- Students should be comfortable reading data and creating basic plots using ggplot2.
- No previous closeread experience is required.
:::

## Data

This module uses a cleaned dataset of winners from the Nathan’s Famous Hot Dog Eating Contest from 1972 to 2025.

Download data: [hotdog_clean.csv]("hotdog_clean.csv")

<details>

<summary><b>Variable Descriptions</b></summary>

| Variable | Description                                               |
|----------|-----------------------------------------------------------|
| Year     | Year the competition took place                           |
| Sex      | Sex of the winner for the year the competition took place |
| Dogs     | Total hot dogs eaten                                      |
| Winner   | Name of the winner for their sex category                 |
| Time     | Time in minutes the competition elapsed                   |
| DPM      | Dogs eaten per minute                                     |

</details>


## Installing closeread

In RStudio:

1. Open the project folder where your .qmd lives.
2. Go to the Terminal tab.
3. Run:

quarto add qmd-lab/closeread


When prompted, type Y to continue.


## Background: Closeread essentials

- Closeread adds one key idea to a normal Quarto document:
- A plot becomes a target when it has an ID like {#cr-plotname}
- Your narrative becomes a trigger when it references the target using @cr-plotname
- When the reader scrolls to that trigger, the plot becomes the focus
- Closeread documents are built with three components:

1) Section

A section enables closeread behavior. Everything inside renders as a scrolling story.

::::{.cr-section}
Narrative + stickies go here
:::

2) Sticky (a target)

A sticky is an element that can be focused (often a plot).

:::{#cr-myplot}
```{r}
# plot code
```
:::
  