---
title: "How Much Does 0.1 Seconds Matter? Time and Ranking in NCAA Men's 50 Freestyle"
author:
  - name: Kevin Euyoque
    email: kevin.euyoque04@gmail.com
    affiliations:
      - ref: slu
  - name: Ivan Ramler
    email: iramler@stlawu.edu
    affiliations:
      - ref: slu
affiliations:
  - id: slu
    name: St. Lawrence University
date: "May 5, 2026"
description: "A data visualization tutorial exploring how small differences in race time relate to rankings in NCAA men's 50 freestyle swimming."
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
categories:
  - Sports Analytics
  - Data Visualization
  - NCAA Swimming
  - R
---

Please note that these material have not yet completed the required pedagogical and industry peer-reviews to become a published module on the SCORE Network. However, instructors are still welcome to use these materials if they are so inclined.

# Welcome Video

If you are unfamiliar with NCAA sprint swimming, watch this short video from the 2024 NCAA Men's 50 Freestyle Championship Final before beginning the module.

<iframe width="560" height="315" src="https://www.youtube.com/embed/jDUU3EdqLrE?si=1-GswmvXKdL2emKT" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>


# Introduction

In competitive swimming, races are often decided by fractions of a second. At the NCAA level, the difference between first place and fifth place can be extremely small, sometimes less than half a second. Despite these tiny margins, the outcomes can be dramatically different.

One of the best examples of this is the **men’s 50 freestyle**, one of the shortest and fastest events in college swimming. Because the race is so short, even a tiny mistake or improvement can noticeably change a swimmer’s finishing position.

In this module, we explore real NCAA swimming data to understand how swim time relates to ranking in the men’s 50 freestyle. Using visualization and linear regression, we investigate how much a small change such as **0.1 seconds** can affect competitive performance.

We will also explore why transforming data using **ln(rank)** improves the regression model and helps satisfy the linearity condition.

::: {.callout-note collapse="true"}
## Learning Objectives

By the end of this module, students will be able to:

1. Understand how performance is measured in competitive swimming  
2. Explore how small differences in swim time impact rank  
3. Analyze the relationship between swim time and rank in NCAA swimming  
4. Interpret a linear regression model using a realistic 0.1-second change  
5. Understand why transformations can improve a regression model  
:::

::: {.callout-note collapse="true"}
## Activity Length

- Approximately 20–30 minutes
:::

::: {.callout-note collapse="true"}
## Methods

**Technology Requirement:**

This activity is designed to be completed in `R` using RStudio or Quarto.

Students should have basic familiarity with:

- reading in a dataset
- using `tidyverse`
- creating scatterplots with `ggplot2`
- interpreting simple linear regression output

No advanced regression experience is required.
:::

## Data

This module uses a cleaned dataset of NCAA swimming performances from the Men’s 50 Freestyle SCY event.

**Swimming dataset**: [SwimData.csv](SwimData.csv)

<details>

<summary><b>Variable Descriptions</b></summary>

| Variable | Description |
|----------|-------------|
| Event | Swimming event name |
| Swimmer | Name of the swimmer |
| Sex | Sex category of the swimmer |
| Swim_Time | Race time measured in seconds |
| Rank | Finishing position in the event |
| ln_rank | Natural logarithm of rank used for regression modeling |

</details>

## Focus Event: Men’s 50 Freestyle

For this analysis, we focus on one event: the **men’s 50 freestyle**.

This event is one of the shortest races in competitive swimming, which makes it ideal for studying how small time differences affect outcomes.

The main variables we will use are:

- **Swim time** (seconds)  
- **Rank** (finishing position)

Lower times are better, and lower ranks are also better.  
So we expect that as swim time increases, rank will also increase.


## Prepare the Data

```{r}
#| echo: TRUE
library(tidyverse)
library(readr)

swim_df <- read_csv("SwimData.csv")

swim_50_free_men <- swim_df |>
  filter(Event == "50 Freestyle SCY Male", Sex == "M") |>
  mutate(Swim_Time = as.numeric(Swim_Time),
         ln_rank = log(Rank)) |>
  select(Event, Swimmer, Sex, Swim_Time, Rank, ln_rank) |>
  drop_na(Swim_Time, Rank, ln_rank)

head(swim_50_free_men)
```

## Visualizing the Relationship

We begin with a scatterplot to explore the relationship between swim time and rank.

```{r}
#| echo: true

ggplot(swim_50_free_men, aes(x = Swim_Time, y = Rank)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  scale_x_continuous(n.breaks = 8) +
  labs(title = "Men's 50 Freestyle: Swim Time vs. Rank",
    subtitle = "The curved pattern suggests the relationship is not perfectly linear",
    x = "Swim Time (seconds)",
    y = "Rank") +
  theme_minimal()
```

This scatterplot shows a clear pattern: as swim time increases, rank also tends to increase.

In swimming, this means that slower times are connected to worse rankings. Since **rank 1 is the best**, moving upward in rank number means finishing lower in the standings.

However, the pattern is slightly curved rather than perfectly linear. This suggests that a simple linear model using raw rank may violate the linearity condition.

Because of this, we transform rank using the natural logarithm.

## Transforming the Relationship

To improve the linear relationship, we model *ln(rank)* instead of rank directly.

Using a logarithmic transformation helps straighten the relationship and makes the regression model more appropriate.

```{r}
#| echo: true

ggplot(swim_50_free_men, aes(x = Swim_Time, y = ln_rank)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(n.breaks = 8) +
  labs(title = "Men's 50 Freestyle: Swim Time vs. ln(Rank)",
    subtitle = "Using ln(rank) helps make the relationship more linear",
    x = "Swim Time (seconds)",
    y = "ln(Rank)") +
  theme_minimal()
```

After transforming rank, the relationship between swim time and performance becomes much more linear.

This is important because linear regression assumes that the response variable changes approximately linearly with the predictor.

The transformed graph shows a strong upward trend:

- faster swim times correspond to better rankings
- slower swim times correspond to worse rankings
- even small changes in time can noticeably affect placement

## Building a Linear Model

Next, we create a linear regression model using swim time to predict rank

```{r}
#| echo: true
rank_model <- lm(ln_rank ~ Swim_Time, data = swim_50_free_men)

summary(rank_model)
```
The regression output shows a strong positive relationship between swim time and *ln(rank)*.

## Interpreting the Model Output

From the regression results, the most important value is the **slope for Swim_Time**, which is:

- **3.151**

Because we modeled *ln(rank)* instead of rank directly, this slope tells us how the logarithm of rank changes as swim time increases.


### What Does the Slope Mean?


- A **1-second increase** in swim time is associated with about a **3.151 increase in ln(rank)**

However, this is not very realistic in this context.

In a 50 freestyle race, a full second is an extremely large difference. NCAA swimmers are often separated by only tenths or hundredths of a second.

So instead, we interpret a 0.1-second increase.

## A More Realistic Interpretation: 0.1 Seconds

Instead, we interpret a **0.1-second change**, which is much more realistic in swimming.

```{r}
#| echo: true
slope <- coef(rank_model)["Swim_Time"]

exp(0.1 * slope)
```

This calculation gives a value of approximately 1.37

This means that a swimmer who is 0.1 seconds slower is predicted to have a rank about 1.37 times worse.

Because lower rank is better in swimming, even a very small increase in time can noticeably hurt a swimmer’s placement.


## Model Strength

The model has an R-squared value of about **0.986**.

This means that about **98.6% of the variation in *ln(rank)*** is explained by swim time.

That is an extremely strong relationship and suggests that swim time is highly connected to ranking in the men’s 50 freestyle.


## Why This Matters

The main takeaway from this analysis is that small differences in time can have a large impact on competitive results.

In the men’s 50 freestyle:

- a 1-second change is unrealistically large
- a 0.1-second change is more realistic
- being 0.1 seconds slower is associated with a substantially worse ranking

This shows why starts, turns, reaction time, and finishing technique matter so much in sprint events. Even a tiny improvement can meaningfully change competitive results.


## Conclusion

In NCAA swimming, fractions of a second are not just small details. They can strongly influence competitive outcomes.

For the men’s 50 freestyle, our model suggests that even a **0.1-second difference** is associated with a meaningful worsening in rank.

By transforming rank using ln(rank), we created a model that better fits the data and more accurately represents the relationship between swim time and performance.

This makes the 50 freestyle a strong example of how statistical modeling can help explain performance differences in sports.

More broadly, this analysis demonstrates how data visualization and regression modeling can be used to better understand competitive performance in athletics.