---
title: "NHL vs KHL: Offense, Defense, and Winning"
format: html
---

# Background

Ice hockey is a team sport where teams try to score goals by shooting a puck into the opponent's net. In this worksheet, each row of the data represents one professional hockey team during the 2024-25 regular season.

We will compare teams from two leagues:

- NHL: National Hockey League, the major professional hockey league in North America.
- KHL: Kontinental Hockey League, a major professional hockey league based mainly in Eurasia.

The main question is:

**Are offense and defense related to winning differently in the NHL and KHL?**

We will compare:

- Win percentage: wins divided by games played.
- Goals for per game: goals scored divided by games played. This measures offense.
- Goals against per game: goals allowed divided by games played. This measures defense. Lower is better.

# 0. Load Packages

```{r}
#| message: false
library(tidyverse)
```

# 1. Load the Data

```{r}
nhl_2024_25 <- read_csv("nhl_20245.csv")
khl_2024_25 <- read_csv("khl_20245.csv")
```

Question: Which columns do we need in order to calculate win percentage, goals for per game, and goals against per game?

# 2. Create Comparable Variables

Create the same five variables for both leagues:

- league
- team
- WinPerc
- GF_pg
- GA_pg

The first dataset is started for you. Complete the KHL code in the same style.

```{r}
nhl_clean <- nhl_2024_25 |>
  mutate(
    league = "NHL",
    team = Team,
    WinPerc = Wins / GamesPlayed,
    GF_pg = GF / GamesPlayed,
    GA_pg = GA / GamesPlayed
  ) |>
  select(league, team, WinPerc, GF_pg, GA_pg)

khl_clean <- khl_2024_25 |>
  mutate(
    league = "KHL",
    team = Team,
    WinPerc = ,
    GF_pg = ,
    GA_pg = 
  ) |>
  select()

hockey <- bind_rows(nhl_clean, khl_clean)
```

Question: Why do we divide by games played instead of just comparing total goals?

# 3. Offense vs Winning

Make a scatterplot showing WinPerc and GF_pg for both leagues. Use facet_wrap() so that each league has its own panel.

```{r}
ggplot(data = hockey, aes(x = , y = )) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~league) +
  labs(
    title = "Offense vs. Winning",
    x = "Goals For per Game",
    y = "Win Percentage")
```

Question: What does this graph suggest about the relationship between offense and winning?

# 4. Defense vs Winning

Make a scatterplot showing WinPerc and GA_pg for both leagues. Use facet_wrap() so that each league has its own panel.

```{r}
ggplot(data = hockey, aes(x = , y = )) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~league) +
  labs(
    title = "Defense vs. Winning",
    x = "Goals Against per Game",
    y = "Win Percentage")
```

Question: What does this graph suggest about the relationship between defense and winning?

# 5. Correlation Summary

Create a table with the offense correlation and defense correlation for each league.

```{r}
hockey |>
  group_by(league) |>
  summarise(
    corr_offense = cor(, , use = "complete.obs"),
    corr_defense = cor(, , use = "complete.obs"))
```

Question: Which correlation is positive? Which correlation is negative? Why does that make sense in hockey terms?


# 6. Reflection Questions

Answer each question in 1-2 sentences.

- Does offense seem related to winning? Use evidence from the graph or correlation table.

- Does defense seem related to winning? Use evidence from the graph or correlation table.

- Which relationship appears stronger in each league: offense with winning, or defense with winning?

- Pretend you are an analyst for a hockey team. Based on this analysis, would you recommend that the team focus more on improving offense, improving defense, or balancing both? Explain your recommendation using the results. 

- Why should we be careful about making a strong conclusion from this analysis?


