---
title: "NHL vs KHL: Offense, Defense, and Winning"
format: html
---

This worksheet compares NHL and KHL teams using one season of data.

The main question is:

**Are offense and defense related to winning differently in the NHL and KHL?**

We will compare:

- goals scored per game
- goals allowed per game
- win percentage

# 0. Load Packages

```{r}
#| message: false
library(tidyverse)
```

# 1. Load the Data

```{r}
nhl_2024_25 <- read_csv("Tim_Kashkarov/offense_defense/nhl_20245.csv")
khl_2024_25 <- read_csv("Tim_Kashkarov/offense_defense/khl_20245.csv")
```

# 2. Create Comparable Variables

Create win percentage, goals scored per game, and goals allowed per game for each league.

```{r}
nhl_2024_25 <- nhl_2024_25 |>
  transmute(
    league = "NHL",
    team = Team,
    WinPerc = Wins / GamesPlayed,
    GF_pg = GF / GamesPlayed,
    GA_pg = GA / GamesPlayed)

khl_2024_25 <- khl_2024_25 |>
  transmute(
    league = "KHL",
    team = Team,
    WinPerc = Wins / GamesPlayed,
    GF_pg = GF / GamesPlayed,
    GA_pg = GA / GamesPlayed)

df <- bind_rows(nhl_2024_25, khl_2024_25)
```

# 3. Offense vs Winning

Plot 2 graphs showing Win Percentage and Goals Scored per Game for both leagues. You may use facet_wrap() to have them look nicer. What does this graph tell us?

```{r}
ggplot(df, aes()) +
  geom_point() +
  geom_smooth() 
```



# 4. Defense vs Winning

Plot 2 graphs showing Win Percentage and Goals Allowed per Game for both leagues. You may use facet_wrap() to have them look nicer. What does this graph tell us?

```{r}
ggplot(df, aes()) +
  geom_point() +
  geom_smooth() 
```



# 5. Correlation Summary

Make offense and defense correlation summary for both leagues. What does this table tell us?  
```{r}

```



# 6. Reflection Questions. Provide answer in 1-2 sentenses.

1. Does offense seem related to winning?



2. Does defense seem related to winning?



3. Which relationship appears stronger in each league?



4. Why should we be careful with this conclusion?


