---
title: "Clash Royale Card Synergy: A Network Analysis Tutorial"
author: "Kevin Euyoque"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(vembedr)
library(igraph)
library(netrankr)
# devtools::install_github("schochastics/networkdata")
```

\*\* Currently working on dataset

Ive made all 125 cards into a dataset however i need to make them into a new one that has the

-   Number of decks with the pair of cards (sample size) DONE
-   Number of wins with the pair of cards 
-   Win rate for those pairs 
-   (any win rate above 50% convert to a 1 and under to a 0)
-   Replace the diagonal of the matrix with zeros 
-   It would be set up with every card listed in the columns and rows with that information for each


We will be focusing on the season 3 of the game.

```{r}
library(readr)
cr_season3_5000decks <- read_csv("cr_season3_5000decks.csv")

cards_id <- pivot_wider(data = cr_season3_5000decks, names_from = card, values_from = winner)

clash_decks <- cards_id |> mutate(across(-id, ~replace(.,!is.na(.),1))) |>
  mutate(across(-id, ~replace(.,is.na(.),0)))

# gets rid of first column
clash_deck_n <- t(as.matrix(clash_decks[,-1])) %*% as.matrix(clash_decks[,-1])


# Clash deck wins
clash_wins <- cards_id |>
  mutate(across(-id, ~ ifelse(is.na(.), 0, as.integer(.))))

clash_deck_wins <- t(as.matrix(clash_wins[, -1])) %*% as.matrix(clash_wins[, -1])


# Clash deck win rate
clash_deck_rate <- clash_deck_wins / clash_deck_n

clash_deck_wins["Zap", "Hog Rider"]
clash_deck_n["Zap", "Hog Rider"]
clash_deck_rate["Zap", "Hog Rider"]
```

```{r}
# Now need to remove the self-pairs
diag(clash_deck_n) <- NA
diag(clash_deck_wins) <- NA
diag(clash_deck_rate) <- NA

```

```{r}
# convert the table into an edge list
pair_data <- as.data.frame(as.table(clash_deck_n)) |>
  rename(card1 = Var1, card2 = Var2, decks = Freq) |>
  mutate(
    wins = as.vector(clash_deck_wins),
    winrate = as.vector(clash_deck_rate)) |>
  filter(card1 != card2)


# Only show decks that have been used more than 50 times
# this allows for only prominently used decks meaning more important
# 1572 rows total
pair_data <- pair_data |>
  filter(decks > 50)

```

```{r}
# any win rate above 50% convert to a 1 and under to a 0
clash_deck_ratebinary <- ifelse(clash_deck_rate > 0.5, 1, 0)

diag(clash_deck_ratebinary) <- 0
```





```{r}
library(tidyverse)

# Shows all the cards available to use 
# including if they have an evo and/or hero upgrade
cards <- read_csv("clash_royale_cards_125.csv")

```


# Introduction: What is Clash Royale?

<iframe width="800" height="450" src="https://www.youtube.com/embed/_hNxfiXmeAE" title="Clash Royale Gameplay" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>

Clash Royale is a real-time strategy mobile game developed by Supercell in which players build decks of eight cards and compete in one on one battles. Each card represents a troop, spell, or building with distinct mechanics and strategic roles.

Success in the game depends heavily on card combinations, as certain cards perform more effectively when used together. Because of this emphasis on synergy, Clash Royale provides a natural setting for studying relationships between elements using network analysis.

Clash Royale was released globally on March 2, 2016

The game has been out for 10 years.

## Goal of this module

In this tutorial, we will treat Clash Royale cards as nodes in a network.

Two cards are connected by an edge if they perform well together, measured by paired win rate.

We will treat:

-   Cards -\> nodes
-   Strong pair win-rates -\> edges
-   Win-rate value -\> edge weight

By the end, we will:

-   Build a synergy network from paired win rates
-   Learn network vocabulary (nodes, edges, weights, degree)
-   Compute centrality (who matters most)
-   Find cliques / communities (tight synergy groups)

# Network Representation

To represent Clash Royale synergy as a network, we will model the data as a graph.

```{r}
# Build the network graph
network <- graph_from_data_frame(
  pair_data,
  directed = FALSE)

network
```
In this we see there are 69 cards that pass the greater than 50 decks filter
- 69 nodes
There is a total of 1572 card pair connections
- 1572 edges


## Nodes

A node represents one Clash Royale card.

Examples of nodes could be:

-   Hog Rider
-   Fireball
-   Miner
-   Ice Spirit

In the network, each card appears once as a node.

## Edges

An **edge** represents a connection between two nodes (two cards).

In this module, an edge means:

The two cards are commonly used together and they perform well together based on paired win rate

So if two cards have strong synergy, they will be connected.

## Edge weights

Edges can also have a weight, which measures how strong the connection is.

For Clash Royale synergy, the edge weight will be the paired win rate for that card pair.

Interpretation:

-   Higher paired win rate → stronger connection
-   Lower paired win rate → weaker connection

```{r}
# Add edge weights
E(network)$weight <- pair_data$winrate

# plot a network
plot(network,
     layout = layout_with_fr,
     vertex.size = 7,
     vertex.label.cex = .7,
     edge.width = E(network)$weight * 5)

# find most connected cards
card_degree <- degree(network)
sort(card_degree, decreasing = TRUE)[1:10]


# Filtered more decks since network too dense
# show only strongest edges
pair_data_small <- pair_data |>
  filter(decks > 150, winrate > 0.52)

network_small <- graph_from_data_frame(pair_data_small, directed = FALSE)

# gives layout more time to spread out
lay <- layout_with_fr(network_small, niter = 5000)

# plot
plot(network_small,
     layout = lay,
     vertex.size = 6,
     vertex.label.cex = 0.75,
     edge.width = E(network_small)$winrate * 4)
```


## Why use a network?

A network is useful here because it lets us study:

-   which cards connect to many others
-   which groups of cards form groups of synergy
-   how the meta might be structured through relationships, not just individual win rates

There are over 37,235,730,424,788 possible unique deck combinations
