---
title: "Closeread on Hotdog Data"
format: closeread-html

---

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
```

```{r}
#| message: false
#| warning: false
hotdog_clean <- read.csv("C:/Users/elena/Downloads/academics/SYE Seminar/stat450-spr2026-score/hotdogs/hotdog_clean.csv")
hotdog_clean_2 <- hotdog_clean |> mutate(standard = Dogs/Time)
```


:::{.cr-section}
The plot looks like this: @cr-plot

The Nathan’s Famous Fourth of July International Hot Dog Eating Contest has a 100+ year tradition. Over time, competitors have pushed the limits of human capacity in ways no one originally imagined.

The Story of Nathan's Hotdog Eating Competition. @cr-time


:::{#cr-plot}

```{r}
ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex), alpha = 0) +
  annotate("text", x = Inf, y = -Inf,
    label = "Years →",
    hjust = 1, vjust = -0.5) +
  annotate("text", x = -Inf, y = Inf,
    label = "↑\nHeavy\npenguins",  #python string notation
    hjust = 0, vjust = 1) +
  guides(colour = "none", shape = "none") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs eaten per minute") 
#  ggsave("filename.png", plot = get_last_plot())
```
:::

:::{#cr-time2}

```{r}
#| message: false
#| warning: false


ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs per minute") +
    geom_vline(xintercept = 2011, linetype = 1) +
      annotate(
            "label",
            x = 2019,
            y = 2.5,
            size = 2.4,
            label = "Female winners") + 
  annotate("rect", xmin = 2010, xmax = 2026, ymin = 2.5, ymax = 5.2,
  alpha = .2, fill = "blue4")
      
```
:::


Early competitions varied in length, which makes raw totals hard to compare across years. To make performances comparable, we standardize results by calculating hot dogs eaten per minute.



This adjustment reveals a much clearer performance trend. Notice how the trajectory of winning totals changes once time is accounted for.



Around the early 2000s, the curve bends sharply upward. This coincides with the arrival of Takeru Kobayashi, whose training methods and eating technique revolutionized the competition.



Beginning in 2011, a new cluster appears: female champions. These competitors show a distinct performance pattern that does not follow the same explosive growth seen in the men’s division.


Focus on 2011–2025

In the most recent years, the darker blue points cluster tightly. This reflects a more stable competitive field and highlights how the women’s division has developed its own trajectory.

