---
title: "Closeread on Hotdog Data"
format: closeread-html

---

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
```

```{r}
#| message: false
#| warning: false
hotdog_clean <- read.csv("C:/Users/elena/Downloads/academics/SYE Seminar/stat450-spr2026-score/hotdogs/hotdog_clean.csv")
```

:::{@cr-sticky}

## The story of Nathan's Hotdog Eating competition
The Nathan’s Famous Fourth of July International Hot Dog Eating Contest has a 100+ year tradition. Here's a story of its iconic winners.

:::

:::{.cr-section}

Closeread enables scrollytelling.

Draw your readers attention with focus effects. @cr-features

:::

:::{.cr-features .aside-right #myMap}
```{r}
#| message: false
#| warning: false
ggplot(hotdog_clean) +
  aes(
    x = Year,
    y = Dogs) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    legend.position.inside = c(0.9, 0.15)) +
  labs(
    x = "Year",
    y = "Dogs eaten")
```
:::




:::{#myid .myclass myattribute="something"}
The curve exhibits some cubic behavior thanks to the introduction of Takeru Kobayashi's new technique based on eating small portions of cabbage and fasting before the competition.
:::



::: {cr-{sticky-D4escription of the need to stadardize}}
The first section of the plot showcases the initial stages of the competition.
Since the number of hot dogs eaten is not standardized for the time, considering that different years had different competition lengths, we need to evaluate a way to standardize all competition times across the years.

Let's compute the ratio of dogs per minute for each year


:::

:::{.cr-section}
```{r}
# Calculating dogs/minute
hotdog_clean_2 <- hotdog_clean |> mutate(standard = Dogs/Time)
```


```{r}
# Updating the plot to reflect the new standardized variable
ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs eaten")

```

The cubic characteristic of the plot is even more emphasized. Why could that possibly be?

### TO-DO highlight The group of female winners that started in 2011. That group of points did not become affected by the standardication...


:::


:::{.cr-section #zoom}

### Focus on 2011–2025

```{r}
ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs eaten")
```

Now we zoom into the recent period where the dark blue points cluster.

:::