---
title: "Closeread on Hotdog Data"
format: closeread-html

---

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
```

```{r}
#| message: false
#| warning: false
hotdog_clean <- read.csv("C:/Users/elena/Downloads/academics/SYE Seminar/stat450-spr2026-score/hotdogs/hotdog_clean.csv")
hotdog_clean_2 <- hotdog_clean |> mutate(standard = Dogs/Time)
```

:::{@cr-sticky}

## The story of Nathan's Hotdog Eating competition
The Nathan’s Famous Fourth of July International Hot Dog Eating Contest has a 100+ year tradition. Here's a story of its iconic winners.

:::

:::{.cr-section}

Closeread enables scrollytelling.

Draw your readers attention with focus effects. @cr-features

:::

:::{.cr-features .aside-right #myMap}
```{r}
#| message: false
#| warning: false
ggplot(hotdog_clean) +
  aes(
    x = Year,
    y = Dogs) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    legend.position.inside = c(0.9, 0.15)) +
  labs(
    x = "Year",
    y = "Dogs eaten")
```
:::
::::: {#special .sidebar}


The curve exhibits some cubic behavior thanks to the introduction of Takeru Kobayashi's new technique based on eating small portions of cabbage and fasting before the competition.
:::



::: {cr-{sticky-D4escription of the need to stadardize}}
The first section of the plot showcases the initial stages of the competition.
Since the number of hot dogs eaten is not standardized for the time, considering that different years had different competition lengths, we need to evaluate a way to standardize all competition times across the years.

Let's compute the ratio of dogs per minute for each year


:::

:::{.cr-section}
```{r}
# Calculating dogs/minute
hotdog_clean_2 <- hotdog_clean |> mutate(standard = Dogs/Time)
```


```{r}
# Updating the plot to reflect the new standardized variable
ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs eaten")

```

The cubic characteristic of the plot is even more emphasized. Why could that possibly be?

### TO-DO highlight The group of female winners that started in 2011. That group of points did not become affected by the standardication...


:::


:::{.cr-section #zoom}

### Focus on 2011–2025


Now we zoom into the recent period where the dark blue points cluster.

:::



-------------------------------
The Story of Nathan's Hotdog Eating Competition

:::{.cr-layout sidebar-left}

:::{.cr-section}
The Nathan’s Famous Fourth of July International Hot Dog Eating Contest has a 100+ year tradition. Over time, competitors have pushed the limits of human capacity in ways no one originally imagined.
:::

:::{.cr-section}
Early competitions varied in length, which makes raw totals hard to compare across years. To make performances comparable, we standardize results by calculating hot dogs eaten per minute.
:::

:::{.cr-section}
This adjustment reveals a much clearer performance trend. Notice how the trajectory of winning totals changes once time is accounted for.
:::

:::{.cr-section}
Around the early 2000s, the curve bends sharply upward. This coincides with the arrival of Takeru Kobayashi, whose training methods and eating technique revolutionized the competition.
:::

:::{.cr-section}
Beginning in 2011, a new cluster appears: female champions. These competitors show a distinct performance pattern that does not follow the same explosive growth seen in the men’s division.
:::

:::{.cr-section}

Focus on 2011–2025

In the most recent years, the darker blue points cluster tightly. This reflects a more stable competitive field and highlights how the women’s division has developed its own trajectory.
:::

:::{.cr-sticky}

```{r}
ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs eaten")
```
:::