---
title: "Hot Dog Closeread Exercise"
subtitle: "Using closeread HTML to tell the story of Nathan's Hot Dog Eating Competition."
description: "A data storytelling module teaching students to use Closeread HTML to visualize data meaningfully."
format:
  closeread-html:
    code-fold: true
    cr-style:
      narrative-background-color-sidebar: "#09753C" # nathan's brand green
      narrative-text-color-sidebar: white
knitr:
  opts_chunk: 
    dev.args:
      bg: transparent
---

### In this exercise, you will create a plot with data from the Wikipedia page for The Story of Nathan's Hot Dog Eating Competition

Loading the necessary libraries to edit the data set and create plots on R.

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
library(plotly)
library(gghighlight)
```

Loading the data set in and performing the ratio calculation to get the dogs eaten per minute.
```{r}
hotdog_clean <- read.csv("C:/Users/elena/Downloads/academics/SYE Seminar/stat450-spr2026-score/hotdogs/hotdog_clean.csv")
hotdog_clean_2 <- hotdog_clean |> mutate(standard = Dogs/Time)

```

Recognize the `sticky` to be included into the closeread document. Rename the sticky to appropriately (though briefly) describe what it is about.

This sticky will appear in within the larger Hot Dog Closeread *scrollytelling* document and it will be referenced within a body of text that describes the plot and its annotations.

Redact a chunk of text that describes the plot you will be creating. This can be edited later to better represent the annotations you produce on the plot.


:::{#cr-}
```{r}
#| label: basicplot
#| message: false
#| warning: false

ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex, size = 1), alpha = 0) +
  annotate("text", x = Inf, y = -Inf,
    label = "Years →",
    hjust = 1, vjust = -0.5) +
  annotate("text", x = -Inf, y = Inf,
    label = "↑\nDogs per\nminute",  #python string notation
    hjust = 0, vjust = 1) +
  guides(colour = "none", shape = "none") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs eaten per minute")
```
:::


