---
title: "Hot Dog Closeread Exercise"
subtitle: "Using closeread HTML to tell the story of Nathan's Hot Dog Eating Competition."
description: "A data storytelling module teaching students to use Closeread HTML to visualize data meaningfully."
format:
  closeread-html:
    code-fold: true
    cr-style:
      narrative-background-color-sidebar: "#09753C" # nathan's brand green
      narrative-text-color-sidebar: white
knitr:
  opts_chunk: 
    dev.args:
      bg: transparent
---

### In this exercise, you will create a plot with data from the Wikipedia page for The Story of Nathan's Hot Dog Eating Competition

Loading the necessary libraries to edit the data set and create plots on R.

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
library(plotly)
library(gghighlight)
```

Loading the data set in and performing the ratio calculation to get the dogs eaten per minute.
```{r}
hotdog_clean <- read.csv("C:/Users/elena/Downloads/academics/SYE Seminar/stat450-spr2026-score/hotdogs/hotdog_clean.csv")
hotdog_clean_2 <- hotdog_clean |> mutate(standard = Dogs/Time)

```

We are first creating a `sticky` to be included into the closeread document. 

0. Rename the sticky  in the R code chunk below to appropriately (though briefly) describe what it is about.

This sticky will appear in within the larger Hot Dog Closeread *scrollytelling* document and it will be referenced within a body of text that describes the plot and its annotations.

1. Redact a brief description for the plot you will be creating. This can be edited later to better represent the annotations you produce on the plot.
| 
| 
| 
| 
| 
| 

2. Edit the R code chunk below to highlight the data points that correspond to Joey Chestnut's wins. Hint: You might find it easier to filter on the data frame itself before editing the plot's aesthetics.
- You can do this by highlighting the data points and changing the shape and color of the points.



3. Add a text label to the plot that describes briefly what the highlighted points represent.
- You could do a `label` or a `text` type annotation.

:::{#cr-}
```{r}
#| label: basicplot
#| echo: false
#| message: false
#| warning: false


joey_chestnut <- hotdog_clean_2 |> 
  
  
ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(aes(colour = factor(Sex))) +
  scale_colour_manual(values = c("#95F527", "#01401C"))+
  annotate("label", x = 1976, y = 6.8,
    label = "Women", fill = "#95F527") +
  annotate("label", x = 1976, y = 7.2,
    label = "Men", fill = "#01401C", color = "#FFFFFF" ) +
  labs(
    x = "Year",
    y = "Dogs eaten per minute") +
      annotate( ) +
  geom_point(data =  , shape =  , color =  , size =  ) +
  theme_minimal() +
  guides(colour = "none", shape = "none") +
  gghighlight::gghighlight(Winner ==  )
```
:::


-------------------

### Alternative setup of the activity (No code provided)

1. Create a plot that highlight the data points that correspond to Joey Chestnut's wins. Hint: You might find it easier to filter on the data frame itself before editing the plot's aesthetics.
- Highlight the data points and change the shape and color of the points.


2. Add a text label to the plot that describes briefly what the highlighted points represent.
- You could do a `label` or a `text` type annotation.




3. Redact a brief description for the plot you created. Mention which specific features of the data are being brought out by the plot and which special R functions were used in the code.




4. Add the reference label for the closeread sticky in the text you redacted to describe the plot. The plot will appear as soon as the reference appears within the mid-line of the screen.s