---
title: "Closeread on Hotdog Data"
format: closeread-html

---

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
```

```{r}
#| message: false
#| warning: false
hotdog_clean <- read.csv("C:/Users/elena/Downloads/academics/SYE Seminar/stat450-spr2026-score/hotdogs/hotdog_clean.csv")
hotdog_clean_2 <- hotdog_clean |> mutate(standard = Dogs/Time)
```


:::: cr-section
The Nathan’s Famous Fourth of July International Hot Dog Eating Contest has a 100+ year tradition. Over time, competitors have pushed the limits of human capacity in ways no one originally imagined.

The Story of Nathan's Hotdog Eating Competition. @cr-time


:::{#cr-time}

```{r}
ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs eaten") 
#  ggsave("filename.png", plot = get_last_plot())
```
:::

:::{#cr-time2}

```{r}
#| message: false
#| warning: false


ggplot(hotdog_clean_2) +
  aes(
    x = Year,
    y = standard) +
  geom_point(
    aes(
      colour = Sex)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Year",
    y = "Dogs per minute") +
    geom_vline(xintercept = 2011, linetype = 1) +
      annotate(
            "label",
            x = 2019,
            y = 2.5,
            size = 2.4,
            label = "Female winners") + 
  annotate("rect", xmin = 2010, xmax = 2026, ymin = 2.5, ymax = 5.2,
  alpha = .2, fill = "blue4")
      
```
:::

:::{.cr-section}
Early competitions varied in length, which makes raw totals hard to compare across years. To make performances comparable, we standardize results by calculating hot dogs eaten per minute.
:::

:::{.cr-section}
This adjustment reveals a much clearer performance trend. Notice how the trajectory of winning totals changes once time is accounted for.
:::

:::{.cr-section}
Around the early 2000s, the curve bends sharply upward. This coincides with the arrival of Takeru Kobayashi, whose training methods and eating technique revolutionized the competition.
:::

:::{.cr-section}
Beginning in 2011, a new cluster appears: female champions. These competitors show a distinct performance pattern that does not follow the same explosive growth seen in the men’s division.
:::

:::{.cr-section}

Focus on 2011–2025

In the most recent years, the darker blue points cluster tightly. This reflects a more stable competitive field and highlights how the women’s division has developed its own trajectory.
:::
